---
title: "Week 04 Lab: Exploring ggplot2 with Pygoscelis Penguins"
subtitle: "Visualizing Penguin Bill Measurements with ggplot2"
author: "Team SLATEBLUE"
date: "2026-05-26"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: false
    theme: cosmo
    code-fold: false
    code-tools: true
    df-print: paged
execute:
  echo: true
  warning: false
  message: false
---


## 3.1 Data Preparation

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(ggbeeswarm)
penguins <- datasets::penguins
```

### Size of the Dataset

```{r}
#| label: data-size
dim(penguins)
nrow(penguins)
ncol(penguins)
```

The `penguins` data frame has **344 rows** and **8 columns**.

---

### Preliminary Exploration

```{r}
#| label: explore-head
head(penguins)
```

```{r}
#| label: explore-tail
tail(penguins)
```

```{r}
#| label: explore-glimpse
glimpse(penguins)
```

**Observations from the output:**

- `species`, `island`, and `sex` are stored as **factors** (`<fct>`), while
  - numeric measurements (`bill_len`, `bill_dep`, `flipper_len`,
  `body_mass`) are stored as **doubles** (`<dbl>`), and `year` as an integer (`<int>`).
- `head()` shows that the first several rows are all Adélie penguins from
  Torgersen island — suggesting the data is **ordered by species**, not
  randomly shuffled.
- `tail()` shows the final rows are Chinstrap penguins from Dream island,
  confirming the species-ordering pattern.
- `glimpse()` reveals **missing values (`NA`)** in several columns,
  particularly `bill_len`, `bill_dep`, and `sex`. Two rows appear to be
  entirely `NA` for the measurement columns, which will need to be removed
  before analysis.
- `body_mass` values are in the thousands (e.g., 3750), so axis labels
  in plots should clearly indicate units (grams).

---

### Data Cleaning

```{r}
#| label: data-cleaning
penguins_bill <- penguins |>
  select(species, bill_len, bill_dep) |>
  drop_na()

nrow(penguins_bill)
```

`penguins_bill` retains the `species`, `bill_len`, and `bill_dep` columns
and removes any rows with missing values in those columns.
After cleaning, **342 rows** remain.

### Verifier Task

```{r}
#| label: verify-data-cleaning
# Count rows removed due to NA in bill_len or bill_dep
rows_removed <- penguins |>
  filter(is.na(bill_len) | is.na(bill_dep)) |>
  nrow()

rows_removed

# Cross-check: original rows minus removed should equal cleaned row count
nrow(penguins) - rows_removed == nrow(penguins_bill)
```

## 3.2 Distribution of Bill Length

### a. Ungrouped Bill-Length Histogram

```{r}
#| label: hist-bill-length-ungrouped
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "Figure 3.2.1: Distribution of bill length across all Pygoscelis penguins. Data: Gorman et al. (2014)."
ggplot(penguins_bill, aes(x = bill_len)) +
  geom_histogram(binwidth = 1, fill = "steelblue", colour = "white") +
  geom_point(
    aes(y = 0),
    position = position_jitter(width = 0.15, height = 0.3),
    alpha = 0.3,
    shape = "|",
    size = 3
  ) +
  labs(
    title = "Distribution of Penguin Bill Length",
    x = "Bill Length (mm)",
    y = "Count"
  ) +
  theme_minimal()
```

### Verifier Task: Histogram Range Check

### Verification

```{r}
#| label: verify-hist-bill-length
# Compute min, max, and range of bill_len
bill_range <- penguins_bill |>
  summarise(
    min_bill = min(bill_len),
    max_bill = max(bill_len),
    range_mm = max_bill - min_bill
  )

bill_range

# Check histogram x-axis covers the full range
# Histogram spans from floor(min) to ceiling(max) with binwidth = 1
expected_min_bin <- floor(bill_range$min_bill)
expected_max_bin <- ceiling(bill_range$max_bill)

cat(
  "Full range that histogram must cover:",
  expected_min_bin,
  "mm to",
  expected_max_bin,
  "mm\n"
)
```

### b. Bin Width Comparison

```{r}
#| label: hist-binwidth-narrow
#| fig-width: 8
#| fig-height: 4
#| fig-cap: "Figure 3.2.2a: Bill length histogram with binwidth = 0.5 mm (narrow). Data: Gorman et al. (2014)."
ggplot(penguins_bill, aes(x = bill_len)) +
  geom_histogram(binwidth = 0.5, fill = "steelblue", colour = "white") +
  labs(
    title = "Penguin Bill Length (binwidth = 0.5 mm)",
    x = "Bill Length (mm)",
    y = "Count"
  ) +
  theme_minimal()
```

```{r}
#| label: hist-binwidth-wide
#| fig-width: 8
#| fig-height: 4
#| fig-cap: "Figure 3.2.2b: Bill length histogram with binwidth = 5 mm (wide). Data: Gorman et al. (2014)."
ggplot(penguins_bill, aes(x = bill_len)) +
  geom_histogram(binwidth = 5, fill = "steelblue", colour = "white") +
  labs(
    title = "Penguin Bill Length (binwidth = 5 mm)",
    x = "Bill Length (mm)",
    y = "Count"
  ) +
  theme_minimal()
```

**Discussion — effect of bin width on apparent distribution shape:**

- **`binwidth = 0.5 mm` (narrow):** The histogram becomes jagged and spiky —
  individual bars represent very small intervals, so random sampling variation
  within each bin dominates the visual. The bimodal shape is still discernible
  but the noise makes it harder to read cleanly. This level of detail is rarely
  useful unless hunting for very fine-grained gaps or outliers in a large dataset.
- **`binwidth = 5 mm` (wide):** The distribution collapses to only about six bars,
  completely smoothing over the bimodal structure. The two humps (around 38–42 mm
  and 46–52 mm) merge into a single broad bar, giving the false impression of a
  roughly unimodal, symmetric distribution. This is a dangerous choice because it
  actively hides the species-level signal present in the data.
- **`binwidth = 1 mm` (chosen in part a):** This is the best balance — the bimodal
  shape is clearly visible, the bars are smooth enough to see structure, and
  individual sampling noise does not dominate. The valley between the two humps at
  roughly 44–45 mm is apparent, which is the key visual cue that two distinct groups
  are pooled together.

**Which bin width to use in a report?** `binwidth = 1 mm` is the most defensible
choice. It honestly represents the shape of the distribution — revealing the
bimodality that motivates the species-level analysis in later sections — without
either obscuring structure (too wide) or introducing visual noise (too narrow).
A good rule of thumb is to choose the smallest bin width at which the overall shape
stabilises and stops changing meaningfully.


### c. Histogram Interpretation

**What explains the non-unimodal structure?**

- **Three species are pooled together:** The `penguins_bill` data combines all
  three Pygoscelis species — Adélie, Gentoo, and Chinstrap — into a single
  histogram without any grouping. If each species has a distinct typical bill
  length, the pooled distribution will naturally show multiple humps, one
  per species cluster.
- **Species differ systematically in bill length:** Adélie penguins tend to
  have shorter bills (roughly 36–46 mm), while Chinstrap and Gentoo penguins
  tend to have longer bills (roughly 44–56 mm). When combined, these two
  clusters produce the two visible humps at approximately 38–42 mm and
  46–52 mm, with a valley around 44–45 mm marking the boundary between them.
- **This is Simpson's Paradox in visual form:** Analysing bill length without
  accounting for species gives a misleading picture of the data's structure.
  A trend or pattern that appears one way in the pooled data may reverse or
  disappear entirely once species is used as a grouping variable — exactly
  the phenomenon this lab is designed to demonstrate.
- **Island and sex may contribute secondary variation:** Since the three
  species are not evenly distributed across Biscoe, Dream, and Torgersen
  islands, and since males tend to have longer bills than females within each
  species, these factors add spread within each hump but are not the primary
  driver of the bimodal shape.

**Implication for the rest of the lab:** The non-unimodal shape is a strong
visual signal that species must be included as a grouping variable in all
subsequent plots. Ignoring it — as the pooled histogram does — risks drawing
conclusions about bill length that are artefacts of mixing distinct biological
groups rather than genuine patterns within any one species.

### d. Raw Point Plot by Species

```{r}
#| label: point-species-bill-length
#| fig-width: 8
#| fig-height: 4
#| fig-cap: "Figure 3.2.4: Bill length by species for Pygoscelis penguins (raw points). Data: Gorman et al. (2014)."
ggplot(penguins_bill, aes(x = bill_len, y = species)) +
  geom_point() +
  labs(
    title = "Bill Length by Species (Raw Points)",
    x = "Bill Length (mm)",
    y = "Species"
  ) +
  theme_minimal()
```

**Observations — apparent density with default `geom_point()`:**

- **Severe overplotting along each row:** Because `species` is a discrete
  factor, all points for a given species are forced onto the same horizontal
  line. Multiple penguins sharing a very similar bill length stack exactly on
  top of one another, rendering as a single point. This makes it impossible
  to judge how many individuals lie at any given x-position.
- **Dense regions appear deceptively sparse:** Areas where many penguins share
  nearly identical bill lengths (e.g., around 38–40 mm for Adélie) look no
  different from areas where only one or two penguins exist — a single solid
  dot is drawn regardless of whether 1 or 15 penguins occupy that position.
- **Species ranges are visible but overlap is unclear:** Despite the
  overplotting, the rough x-axis range for each species can be read — Adélie
  points cluster at shorter lengths, while Gentoo points extend further right.
  However, the degree of overlap between Chinstrap and Adélie near 44–46 mm
  cannot be reliably assessed from this plot alone.
- **Implication:** The default `geom_point()` is unsuitable here because the
  discrete y-axis guarantees overplotting. Mitigation strategies such as
  `geom_jitter()`, `geom_count()`, or `ggbeeswarm::geom_beeswarm()` are
  needed to reveal the true density of observations along each species row.


  ### e. Jitter Plot by Species

```{r}
#| label: jitter-species-bill-length
#| fig-width: 8
#| fig-height: 4
#| fig-cap: "Figure 3.2.5: Bill length by species with vertical jitter (geom_jitter()). Data: Gorman et al. (2014)."
ggplot(penguins_bill, aes(x = bill_len, y = species)) +
  geom_jitter(
    width = 0,
    height = 0.2,
    alpha = 0.4,
    size = 1.5
  ) +
  labs(
    title = "Bill Length by Species (Jittered Points)",
    x = "Bill Length (mm)",
    y = "Species"
  ) +
  theme_minimal()
```

**What jittering reveals that raw points could not:**

- **True density is now visible within each species:** By spreading points
  vertically, overlapping observations no longer collapse into a single dot.
  Regions where many penguins share similar bill lengths now show a visibly
  denser cluster of points, while sparse regions remain visually open —
  giving an honest impression of the local frequency of observations.
- **Adélie has a compact, narrowly concentrated distribution:** The Adélie
  points cluster tightly between roughly 36–46 mm, with the densest band
  around 38–42 mm. There are very few Adélie penguins with bill lengths
  above 46 mm, confirming a clear upper boundary for this species.
- **Chinstrap and Gentoo occupy higher but distinct ranges:** Chinstrap
  points concentrate around 46–52 mm, while Gentoo points span a wider
  range extending toward 60 mm. The two species do overlap somewhat, but
  their density peaks are visibly offset from each other.
- **The valley in the pooled histogram is explained:** The gap around
  44–45 mm in the Section 3.2a histogram corresponds to the boundary between
  the Adélie cluster and the Chinstrap/Gentoo clusters — now clearly visible
  as a low-density zone between species rows.
- **`width = 0` is essential for x-axis integrity:** Setting horizontal
  jitter to zero ensures that each point's x-position accurately reflects
  the penguin's actual bill length. Introducing horizontal jitter would
  misrepresent the measurement values and undermine any x-axis reading.
- **`height = 0.2` keeps species rows separated:** A vertical jitter of 0.2
  is sufficient to spread overlapping points without causing points from
  adjacent species (which are 1.0 unit apart on the y-axis) to intermingle,
  preserving the visual association between each point and its species.

  ### f. Beeswarm Plot by Species

```{r}
#| label: beeswarm-species-bill-length
#| fig-width: 8
#| fig-height: 4
#| fig-cap: "Figure 3.2.6: Bill length by species using a beeswarm plot (geom_beeswarm()). Data: Gorman et al. (2014)."
ggplot(penguins_bill, aes(x = bill_len, y = species)) +
  geom_beeswarm(
    cex = 1.8,
    alpha = 0.6,
    size = 1.5
  ) +
  labs(
    title = "Bill Length by Species (Beeswarm)",
    x = "Bill Length (mm)",
    y = "Species"
  ) +
  theme_minimal()
```

**How the beeswarm plot compares to the jitter plot:**

- **Systematic vs. random offsetting:** `geom_jitter()` displaces points by
  a random vertical amount within the specified `height` range, so the layout
  changes every time the plot is rendered (unless a seed is set). 
  `geom_beeswarm()` uses a deterministic algorithm that offsets each point
  just enough to avoid overlap with its neighbours, producing a stable and
  reproducible layout.
- **Every point stays close to its true x-value:** Because the beeswarm
  algorithm only offsets points perpendicular to the data axis (vertically
  here), each point remains at its exact measured bill length on the x-axis.
  No horizontal noise is introduced, making x-axis readings more trustworthy
  than in a jitter plot where accidental horizontal drift can occur if
  `width` is not explicitly set to zero.
- **Density is encoded in the width of the swarm:** Where many penguins share
  similar bill lengths, the swarm fans out vertically to accommodate all
  points — creating a wider band at dense x-positions and a narrow band at
  sparse ones. This gives the beeswarm an implicit histogram-like shape,
  communicating distributional density without binning.
- **Adélie's compact cluster and Gentoo's spread are more legible:** The
  beeswarm makes it easier to see that Adélie penguins form a tight, tall
  swarm around 38–42 mm (many individuals at similar lengths), while Gentoo
  penguins spread into a flatter, wider swarm across a broader range.
- **Trade-off — beeswarm requires more vertical space:** For large datasets,
  the swarm can become very tall and may cause species rows to visually
  overlap if `fig-height` is insufficient. The jitter plot is less sensitive
  to this because random vertical displacement is bounded by `height`.
- **Preferred choice for a report:** The beeswarm is the stronger choice
  when the goal is to communicate both individual data points and distributional
  shape simultaneously, as its layout is reproducible and its density encoding
  is direct. The jitter plot is acceptable as a quick exploratory view but is
  less reliable for a polished report due to its random layout.

  ### g. Pros and Cons — Jitter vs. Beeswarm

- **Advantage of beeswarm over jitter:**
The beeswarm plot encodes local density directly through the visible width of
the swarm at each x-position. For example, the dense cluster of Adélie penguins
around 38–42 mm fans out into a noticeably taller column of points, making it
immediately clear that many individuals share those bill lengths. In the jitter
plot, the same region appears as a rough band of randomly scattered dots —
the density signal is present but less structured, and a reader must mentally
integrate the dot cloud rather than reading density from a clean visual shape.

- **Disadvantage of beeswarm over jitter:**
The beeswarm layout requires more vertical space per species row because every
point must be physically offset to avoid overlap. With 342 observations across
three species, the Gentoo and Chinstrap swarms in particular fan out to a
considerable height. If the figure dimensions are not generous enough, adjacent
species rows can visually crowd each other, making it harder to associate a
point with its correct species. The jitter plot, with its bounded `height`
parameter, keeps each species row within a fixed vertical band regardless of
how many points cluster at a given x-position — making it more robust to
crowding at small figure sizes.


### h. Faceted Bill-Length Histogram by Species

```{r}
#| label: hist-faceted-species
#| fig-width: 10
#| fig-height: 6
#| fig-cap: "Figure 3.2.8: Bill length distribution faceted by species. Dashed vertical line marks the overall mean bill length across all species. Data: Gorman et al. (2014)."
# Compute overall mean once — scalar, so geom_vline() places it identically
# in every facet
overall_mean_bill <- mean(penguins_bill$bill_len)

ggplot(penguins_bill, aes(x = bill_len)) +
  geom_histogram(binwidth = 1, fill = "steelblue", colour = "white") +
  geom_point(
    aes(y = 0),
    position = position_jitter(width = 0.15, height = 0.3),
    alpha = 0.3,
    shape = "|",
    size = 3
  ) +
  geom_vline(
    xintercept = overall_mean_bill,
    colour = "firebrick",
    linetype = "dashed",
    linewidth = 0.8
  ) +
  facet_wrap(~species, ncol = 1) +
  labs(
    title = "Bill Length Distribution by Species",
    subtitle = paste0(
      "Dashed line = overall mean (",
      round(overall_mean_bill, 1),
      " mm)"
    ),
    x = "Bill Length (mm)",
    y = "Count"
  ) +
  theme_minimal()
```

**Observations from the faceted histogram:**

- **Each species is unimodal in isolation:** Once the data are split by
  species, each facet shows a clean single-peaked distribution. The bimodal
  shape seen in the pooled histogram in Section 3.2a is entirely explained
  by mixing three distinct species — not by any within-species variation.
- **Adélie penguins fall entirely below the overall mean:** The overall mean
  (~43.9 mm) sits to the right of the Adélie distribution's peak (~39 mm),
  meaning nearly all Adélie individuals have shorter bills than the
  cross-species average. The anchor line makes this species-level offset
  immediately visible without any calculation.
- **Chinstrap penguins straddle the overall mean:** The Chinstrap distribution
  is centred close to the overall mean line, with roughly equal numbers of
  individuals on either side — reflecting that their bill lengths happen to
  be close to the pooled average.
- **Gentoo penguins fall almost entirely above the overall mean:** The Gentoo
  distribution peaks well to the right of the anchor line (~47–50 mm),
  confirming that Gentoo penguins consistently have longer bills than the
  cross-species average.
- **The jittered rug confirms within-facet individual spread:** The rug layer
  draws only from the data in each facet (ggplot2 automatically subsets data
  per facet), so the tick marks correctly represent only the individuals in
  that species row — no cross-species contamination occurs.


### i. Violin-with-Box Plot

```{r}
#| label: violin-box-species-bill-length
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "Figure 3.2.9: Violin plot with embedded box plot of bill length by species. Data: Gorman et al. (2014)."
ggplot(penguins_bill, aes(x = bill_len, y = species)) +
  geom_violin(fill = "steelblue", alpha = 0.4, colour = "grey40") +
  geom_boxplot(width = 0.08, outlier.shape = NA, colour = "grey20") +
  labs(
    title = "Bill Length by Species (Violin + Box Plot)",
    x = "Bill Length (mm)",
    y = "Species"
  ) +
  theme_minimal()
```

**Observations from the violin-with-box plot:**

- **Distributional shape and summary statistics in one view:** The violin
  layer shows the full density estimate for each species — where the violin
  is widest, observations are most concentrated. The embedded box plot adds
  the median, interquartile range (IQR), and whiskers on top of that shape,
  providing precise summary statistics without a separate plot.
- **Adélie has the narrowest, most symmetric violin:** The Adélie violin is
  compact and roughly symmetric around its median (~39 mm), confirming the
  tight, unimodal distribution seen in the faceted histogram. The IQR box
  is short, indicating low spread among the middle 50% of individuals.
- **Chinstrap has a tall, narrow violin centred higher:** The Chinstrap
  violin peaks around 49–50 mm with relatively little spread, suggesting
  this species has the most consistent bill lengths of the three. The median
  sits clearly above the overall cross-species range of Adélie.
- **Gentoo has the widest and most spread-out violin:** The Gentoo violin is
  broader and extends further right than the other two species, reflecting
  greater within-species variability in bill length. The longer whiskers and
  wider IQR box confirm this spread.
- **`outlier.shape = NA` avoids double-plotting:** Since the violin already
  shows the full distribution including extremes, suppressing box plot
  outlier points prevents them from being plotted twice and cluttering the
  figure. The violin's tails communicate the same information more smoothly.
- **`width = 0.08` keeps the box plot narrow:** A thin box plot sits cleanly
  inside the violin without obscuring the density shape, preserving the
  visual hierarchy — distribution shape first, summary statistics second.

### j. Why Combine Violin and Box Plots

A box plot alone reduces each distribution to five summary statistics —
minimum, Q1, median, Q3, and maximum — which cannot distinguish between
a unimodal bell-shaped distribution and a bimodal or skewed one with the
same quartile values. A violin plot alone shows the full density shape but
provides no precise numerical anchors, making it difficult to compare
medians or IQRs across species at a glance. Together, the violin supplies
the distributional context that the box plot suppresses, while the box plot
supplies the exact summary statistics that the violin only implies — so a
reader can simultaneously ask *"what shape is this distribution?"* and
*"where exactly is the median and how wide is the middle 50%?"* without
switching between two separate plots.


## 3.3 Distribution of Bill Depth

### a. Ungrouped Bill-Depth Histogram

```{r}
#| label: hist-bill-depth-ungrouped
#| fig-width: 8
#| fig-height: 5
#| fig-cap: "Figure 3.3.1: Distribution of bill depth across all Pygoscelis penguins. Data: Gorman et al. (2014)."
ggplot(penguins_bill, aes(x = bill_dep)) +
  geom_histogram(binwidth = 0.5, fill = "coral", colour = "white") +
  geom_point(
    aes(y = 0),
    position = position_jitter(width = 0.1, height = 0.3),
    alpha = 0.3,
    shape = "|",
    size = 3
  ) +
  labs(
    title = "Distribution of Penguin Bill Depth",
    x = "Bill Depth (mm)",
    y = "Count"
  ) +
  theme_minimal()
```

**How bill depth compares to bill length:**

- **Bill depth is also non-unimodal when pooled:** Like bill length, the
  pooled bill-depth histogram does not show a single clean peak. There is
  a broader spread with possible shoulders or secondary humps, again
  suggesting that mixing three species with different typical bill depths
  produces a multi-modal appearance in the aggregate.
- **Bill depth spans a narrower absolute range:** Bill length ranged from
  approximately 32–60 mm (~28 mm range), while bill depth spans roughly
  13–21 mm (~8 mm range). The distribution is therefore more compressed
  on the x-axis, and a smaller `binwidth = 0.5` mm is appropriate to
  reveal within-species structure without over-smoothing.
- **The two humps are less clearly separated than in bill length:** In the
  bill-length histogram, the valley around 44–45 mm was a prominent gap
  between the Adélie cluster and the Chinstrap/Gentoo clusters. In the
  bill-depth histogram, the species clusters overlap more, making the
  pooled distribution appear less distinctly bimodal and more like a
  broad, flat-topped single peak — suggesting bill depth alone may be
  a weaker species separator than bill length when viewed in isolation.
- **The rug confirms no large isolated outliers:** The tick marks along
  the bottom edge are distributed continuously across the full range with
  no obvious isolated stragglers far from the main body, indicating that
  the spread in bill depth reflects genuine species-level variation rather
  than measurement anomalies.


### b. Faceted Bill-Depth Histogram by Species

```{r}
#| label: hist-faceted-bill-depth
#| fig-width: 10
#| fig-height: 6
#| fig-cap: "Figure 3.3.2: Bill depth distribution faceted by species. Dashed vertical line marks the overall mean bill depth across all species. Data: Gorman et al. (2014)."
overall_mean_depth <- mean(penguins_bill$bill_dep)

ggplot(penguins_bill, aes(x = bill_dep)) +
  geom_histogram(binwidth = 0.5, fill = "coral", colour = "white") +
  geom_point(
    aes(y = 0),
    position = position_jitter(width = 0.1, height = 0.3),
    alpha = 0.3,
    shape = "|",
    size = 3
  ) +
  geom_vline(
    xintercept = overall_mean_depth,
    colour = "firebrick",
    linetype = "dashed",
    linewidth = 0.8
  ) +
  facet_wrap(~species, ncol = 1) +
  labs(
    title = "Bill Depth Distribution by Species",
    subtitle = paste0(
      "Dashed line = overall mean (",
      round(overall_mean_depth, 1),
      " mm)"
    ),
    x = "Bill Depth (mm)",
    y = "Count"
  ) +
  theme_minimal()
```

**How bill depth differs across species:**

- **Each species is unimodal in isolation, as with bill length:** Once
  faceted, the broad and ambiguous pooled distribution resolves into three
  clean single-peaked distributions per species — confirming again that
  the pooled non-unimodality is entirely a consequence of mixing species,
  not within-species variation.
- **Gentoo penguins have markedly shallower bills:** The Gentoo facet peaks
  around 14–15 mm and sits almost entirely to the left of the overall mean
  (~17.2 mm). This is the most striking species-level contrast — Gentoo
  penguins have both the longest and the shallowest bills of the three
  species, making bill depth a complementary separator to bill length.
- **Adélie and Chinstrap have similar, deeper bills:** Both Adélie and
  Chinstrap distributions are centred above the overall mean, with peaks
  around 18–19 mm. Their bill-depth distributions overlap considerably,
  meaning bill depth alone cannot reliably separate these two species from
  each other — unlike bill length, where they occupy more distinct ranges.
- **Bill depth and bill length provide complementary information:** Bill
  length separates Adélie from the other two species most clearly, while
  bill depth separates Gentoo from the other two most clearly. Neither
  dimension alone perfectly distinguishes all three species, which
  motivates combining both measurements in a two-dimensional scatter plot
  in Section 3.4.
- **The anchor line highlights the Gentoo offset most dramatically:** The
  overall mean depth line falls to the right of the entire Gentoo
  distribution, visually reinforcing how far below average Gentoo bill
  depths are — an effect that was completely invisible in the pooled
  histogram.

### c. Comparison of Bill-Length and Bill-Depth Distributions

**Which species does each measurement separate?**

- **Bill length (Section 3.2e–j) separates Adélie from Chinstrap and
  Gentoo:** Across the jitter plot, beeswarm, faceted histogram, and
  violin-box plot, Adélie penguins consistently occupy a shorter bill-length
  range (~36–46 mm) that is clearly offset from both Chinstrap (~46–56 mm)
  and Gentoo (~42–60 mm). However, Chinstrap and Gentoo distributions
  overlap substantially in bill length, meaning bill length alone cannot
  reliably distinguish those two species from each other.
- **Bill depth (Section 3.3b) separates Gentoo from Adélie and Chinstrap:**
  The faceted histogram shows Gentoo penguins clustered at shallower depths
  (~13–17 mm), well below the overall mean, while Adélie and Chinstrap both
  peak around 18–19 mm and overlap heavily with each other. Bill depth
  therefore draws a clear boundary around Gentoo but leaves Adélie and
  Chinstrap indistinguishable on this dimension alone.
- **The two measurements separate different pairs of species:** Bill length
  isolates Adélie as the short-billed outlier; bill depth isolates Gentoo
  as the shallow-billed outlier. Critically, neither measurement on its own
  can separate all three species simultaneously — the pair that overlaps in
  one dimension is the pair that is well-separated in the other.
- **Combining both measurements is necessary for full three-way separation:**
  Since bill length separates {Adélie} from {Chinstrap, Gentoo} and bill
  depth separates {Gentoo} from {Adélie, Chinstrap}, using both dimensions
  together should allow all three species to occupy distinct regions of a
  two-dimensional plot — which the scatter plot in Section 3.4 will test
  directly.


### Verifier Task — Mean Bill Dimensions by Species

```{r}
#| label: verify-bill-means-by-species
# Compute mean bill length and depth per species
bill_means <- penguins_bill |>
  summarise(
    mean_bill_len = round(mean(bill_len), 2),
    mean_bill_dep = round(mean(bill_dep), 2),
    n = n(),
    .by = species
  ) |>
  arrange(mean_bill_len)

bill_means
```

```{r}
#| label: verify-overlap-check
# Check pairwise differences in mean bill length and depth
bill_means |>
  summarise(
    adelie_chinstrap_length_diff = mean_bill_len[species == "Chinstrap"] -
      mean_bill_len[species == "Adelie"],
    chinstrap_gentoo_length_diff = mean_bill_len[species == "Gentoo"] -
      mean_bill_len[species == "Chinstrap"],
    adelie_chinstrap_depth_diff = mean_bill_dep[species == "Adelie"] -
      mean_bill_dep[species == "Chinstrap"],
    gentoo_adelie_depth_diff = mean_bill_dep[species == "Adelie"] -
      mean_bill_dep[species == "Gentoo"]
  )
```

**Interpreting the verification output:**

- **Bill length means confirm the Adélie separation:** Adélie has the
  shortest mean bill length (~38.8 mm), sitting well below both Chinstrap
  (~48.8 mm) and Gentoo (~47.5 mm), consistent with the clear left-offset
  seen in the jitter and beeswarm plots. The ~10 mm gap between Adélie and
  the other two species is large enough to justify the visual claim that
  bill length separates Adélie cleanly.
- **Chinstrap and Gentoo means are close in bill length:** The difference
  between Chinstrap and Gentoo mean bill lengths is small (~1–2 mm),
  numerically confirming the visual overlap observed in Section 3.2 —
  bill length alone cannot separate these two species reliably.
- **Bill depth means confirm the Gentoo separation:** Gentoo has the
  shallowest mean bill depth (~15.0 mm), roughly 3–4 mm below both Adélie
  (~18.3 mm) and Chinstrap (~18.4 mm). This large gap corroborates the
  visual claim that bill depth isolates Gentoo from the other two species.
- **Adélie and Chinstrap depth means are nearly identical:** The near-zero
  difference between their mean bill depths numerically validates the
  visual finding that these two species are indistinguishable on the
  bill-depth dimension alone.


## 3.4 Bivariate View: Bill Depth versus Bill Length

### a. Ungrouped Scatter Plot with Pooled Linear Trend

```{r}
#| label: scatter-ungrouped-bill
#| fig-width: 8
#| fig-height: 6
#| fig-cap: "Figure 3.4.1: Scatter plot of bill depth vs. bill length across all Pygoscelis penguins with pooled linear trend. Data: Gorman et al. (2014)."
ggplot(penguins_bill, aes(x = bill_len, y = bill_dep)) +
  geom_point(alpha = 0.4, size = 1.8) +
  geom_smooth(method = "lm", colour = "firebrick", se = TRUE) +
  labs(
    title = "Bill Depth vs. Bill Length (Pooled, All Species)",
    x = "Bill Length (mm)",
    y = "Bill Depth (mm)"
  ) +
  theme_minimal()
```

**Observations from the ungrouped scatter plot:**

- **The pooled trend is negative:** The fitted regression line slopes
  downward from left to right, suggesting that — when all species are
  pooled together — penguins with longer bills tend to have shallower
  bills. A naive reading of this plot would lead to the conclusion that
  bill length and bill depth are negatively correlated.
- **The point cloud does not support a single coherent linear relationship:**
  Rather than a uniform scatter around the trend line, the points form two
  or three loosely separated clusters. The pooled regression line cuts
  through the gaps between these clusters rather than through the centre
  of any one of them, suggesting the line is not a meaningful description
  of the relationship within any subgroup.
- **This is a textbook example of Simpson's Paradox:** The negative pooled
  trend is driven entirely by the compositional mix of species — Adélie
  penguins (short bills, deep bills) pull the left side of the line up,
  while Gentoo penguins (long bills, shallow bills) pull the right side
  down. The aggregate trend is an artefact of combining groups with
  different means, not a genuine within-species relationship.
- **Species must be mapped to colour to interpret this plot correctly:**
  Without species information, the scatter plot is actively misleading.
  The next step — colouring points by species and fitting per-species
  trend lines — is necessary to reveal whether bill length and bill depth
  are positively or negatively related within each species individually.


### b. Interpretation of the Pooled Trend

Without species information, a reader would infer a **negative relationship**
between bill length and bill depth — that is, penguins with longer bills tend
to have shallower bills, with the regression line sloping visibly downward
across the full range of bill lengths (~32–60 mm). The confidence band around
the pooled trend line is narrow throughout and does not widen to overlap a
horizontal line at any point along the x-axis, indicating that the negative
slope is statistically significant and unlikely to be due to chance alone.
However, this conclusion is entirely misleading — the negative trend is an
artefact of Simpson's Paradox, where the compositional mix of three species
with different bill-length and bill-depth means produces a spurious aggregate
slope that reverses or disappears once species is accounted for.

### c. Exact Overplotting

```{r}
#| label: verify-exact-overplotting
# Count observations at each unique (bill_len, bill_dep) position
overlap_counts <- penguins_bill |>
  count(bill_len, bill_dep, name = "stack_size")

# Number of points hidden by exact overlap
# For each position with n > 1, (n - 1) points are hidden behind the top point
points_hidden <- overlap_counts |>
  summarise(hidden = sum(stack_size - 1)) |>
  pull(hidden)

# Maximum stack size at any single (x, y) position
max_stack <- max(overlap_counts$stack_size)

cat("Points hidden by exact overlap:", points_hidden, "\n")
cat("Maximum stack size at any (x, y) position:", max_stack, "\n")
```

**Discussion — is exact coincidence the only cause of overplotting?**

- **Exact coincidence accounts for only a small fraction of overplotting:**
  The two numbers above quantify only cases where two or more penguins share
  precisely the same bill length and bill depth values to the nearest 0.1 mm.
  In practice this is relatively rare, since continuous measurements recorded
  to one decimal place have limited repetition.
- **Near-coincidence is the dominant overplotting problem here:** Even when
  two points do not share identical coordinates, points that are very close
  together — say, 0.1–0.3 mm apart in both dimensions — will visually overlap
  at typical figure sizes. The human eye cannot resolve two dots whose centres
  are only a fraction of a millimetre apart on a printed or screen plot,
  making them perceptually indistinguishable from exact duplicates.
- **Cluster-level crowding compounds the problem:** The three species occupy
  distinct but internally dense regions of the bill-length × bill-depth space.
  Within each cluster, many penguins share similar measurements, producing
  regions of heavy dot density where individual points cannot be counted or
  separated regardless of whether they are exactly coincident. Techniques such
  as `alpha` transparency, `geom_count()`, or `geom_jitter()` are needed to
  communicate this density honestly.

### Verifier Task — Find hidden points using distinct()

```{r}
#| label: verify-overplotting-distinct
# Total rows minus distinct (x, y) positions = total hidden points
total_rows <- nrow(penguins_bill)
distinct_positions <- penguins_bill |>
  distinct(bill_len, bill_dep) |>
  nrow()

points_hidden_distinct <- total_rows - distinct_positions

cat("Total observations:        ", total_rows, "\n")
cat("Distinct (x, y) positions: ", distinct_positions, "\n")
cat("Points hidden by overlap:  ", points_hidden_distinct, "\n")
```


### d. Mitigation of Overplotting

#### Variant 1: Jittered Scatter Plot

```{r}
#| label: scatter-jitter-bill
#| fig-width: 8
#| fig-height: 6
#| fig-cap: "Figure 3.4.4a: Jittered scatter plot of bill depth vs. bill length with pooled linear trend. Data: Gorman et al. (2014)."
ggplot(penguins_bill, aes(x = bill_len, y = bill_dep)) +
  geom_jitter(
    width = 0.15,
    height = 0.08,
    alpha = 0.4,
    size = 1.8
  ) +
  geom_smooth(method = "lm", colour = "firebrick", se = TRUE) +
  labs(
    title = "Bill Depth vs. Bill Length (Jittered + Pooled Trend)",
    x = "Bill Length (mm)",
    y = "Bill Depth (mm)"
  ) +
  theme_minimal()
```

**Observations (jittered scatter plot):**

- Jittering separates points that were previously stacked at identical or
  near-identical coordinates, making the three loose clusters more visually
  distinct and allowing local density to be perceived through dot
  concentration rather than dot darkness alone.
- `width = 0.15` and `height = 0.08` are kept small relative to the
  measurement scale so that displacement does not meaningfully misrepresent
  any individual penguin's actual bill dimensions — positional accuracy is
  slightly sacrificed but not abandoned.
- The pooled regression line remains negative and nearly identical to that
  in Section 3.4a, confirming that jittering does not alter the fitted
  trend — it only affects point placement, not the underlying data passed
  to `geom_smooth()`.

---

#### Variant 2: Count-Symbol Plot

```{r}
#| label: scatter-count-bill
#| fig-width: 8
#| fig-height: 6
#| fig-cap: "Figure 3.4.4b: Count-symbol plot of bill depth vs. bill length. Symbol area is proportional to the number of penguins at each coordinate. Pooled linear trend included. Data: Gorman et al. (2014)."
ggplot(penguins_bill, aes(x = bill_len, y = bill_dep)) +
  geom_count(alpha = 0.5, colour = "steelblue") +
  scale_size_area(
    name = "Count",
    max_size = 10
  ) +
  geom_smooth(method = "lm", colour = "firebrick", se = TRUE) +
  labs(
    title = "Bill Depth vs. Bill Length (Count Symbols + Pooled Trend)",
    x = "Bill Length (mm)",
    y = "Bill Depth (mm)"
  ) +
  theme_minimal()
```

**Observations (count-symbol plot):**

- `geom_count()` places a single symbol at each unique (x, y) coordinate
  and scales its area proportionally to the number of observations stacked
  there. Points that appeared identical in the raw scatter plot are now
  represented by visibly larger circles, making exact overplotting
  immediately legible without any positional displacement.
- `scale_size_area()` ensures that symbol **area** — not radius — is
  proportional to count, which is the perceptually correct encoding for
  quantity. Doubling the count produces a symbol that appears twice as
  large in area, not twice as wide in radius.
- Unlike jittering, `geom_count()` preserves the exact measured coordinates
  of every point — no positional noise is introduced — making it more
  faithful to the raw data for readers who want to read off precise values.
- The trade-off is that near-coincident points (within 0.1 mm of each other
  but not exactly equal) are not merged by `geom_count()` and may still
  overlap visually, whereas jittering disperses them regardless of whether
  they are exactly coincident.


### e. Assessment of Scatter Plot Variants

For a journal submission, the **count-symbol plot (Section 3.4d.ii)** is the
strongest choice. It preserves the exact measured coordinates of every penguin
— unlike the jittered variant, which introduces artificial positional noise
that could mislead a reader attempting to read off precise bill dimensions —
while making the degree of overplotting explicitly legible through symbol area.
The `scale_size_area()` encoding is perceptually honest, as area is
proportional to count, so a referee can immediately assess where observations
are concentrated without mentally integrating a cloud of dots as in variants
(a) and (d.i). The pooled regression line is also more interpretable against
the count-symbol background because the fitted trend is visually anchored to
circles whose sizes reflect true data density, making it clearer which regions
of the plot are driving the slope.

### f. Grouped Scatter Plot by Species

```{r}
#| label: scatter-grouped-species
#| fig-width: 9
#| fig-height: 6
#| fig-cap: "Figure 3.4.6: Count-symbol scatter plot of bill depth vs. bill length, coloured by species with per-species linear trends. Data: Gorman et al. (2014)."
ggplot(
  penguins_bill,
  aes(x = bill_len, y = bill_dep, colour = species)
) +
  geom_count(alpha = 0.6) +
  scale_size_area(name = "Count", max_size = 10) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "Bill Depth vs. Bill Length by Species",
    x = "Bill Length (mm)",
    y = "Bill Depth (mm)",
    colour = "Species"
  ) +
  theme_minimal()
```

**What becomes visible when species is mapped to colour:**

- **All three within-species trends are positive:** Once species is
  distinguished by colour, the regression line for each species slopes
  upward — penguins with longer bills within the same species also tend
  to have deeper bills. This is the exact opposite direction to the
  pooled trend in Section 3.4a, which sloped downward. The negative
  pooled trend was entirely an artefact of mixing three groups with
  different means, not a genuine biological relationship.
- **Three distinct clusters become immediately apparent:** The coloured
  points reveal that what appeared to be a single diffuse cloud in the
  ungrouped plot is actually three well-separated clusters — Adélie
  (short, deep bills), Chinstrap (long, deep bills), and Gentoo (long,
  shallow bills). The overlap between clusters is minimal, confirming
  that the combination of both bill dimensions separates all three
  species far more cleanly than either dimension alone.
- **Gentoo's positive within-species slope is the steepest:** The Gentoo
  regression line rises more steeply than those of Adélie and Chinstrap,
  suggesting that bill length is a stronger predictor of bill depth
  within Gentoo penguins than within the other two species.

**Statistical term for the paradox:**

The apparent reversal of the trend direction — negative when pooled across
all species, positive within each species individually — is a classic
instance of **Simpson's Paradox**: an aggregate trend across grouped data
that reverses or disappears when the groups are analysed separately. It
arises here because species acts as a confounding variable; the three
species differ systematically in both bill length and bill depth, so
pooling them conflates between-group differences with within-group
relationships and produces a spurious negative correlation.


### Verifier Task — Pearson Correlation by Species

```{r}
#| label: verify-correlation-pooled
# Pooled correlation across all species
pooled_cor <- cor(penguins_bill$bill_len, penguins_bill$bill_dep)

cat("Pooled Pearson correlation (all species):", round(pooled_cor, 4), "\n")
cat(
  "Sign:",
  ifelse(
    pooled_cor < 0,
    "NEGATIVE — matches downward pooled trend\n",
    "POSITIVE\n"
  )
)
```

```{r}
#| label: verify-correlation-by-species
# Per-species Pearson correlation
species_cor <- penguins_bill |>
  summarise(
    correlation = round(cor(bill_len, bill_dep), 4),
    sign = ifelse(
      cor(bill_len, bill_dep) > 0,
      "positive",
      "negative"
    ),
    n = n(),
    .by = species
  ) |>
  arrange(species)

species_cor

# Verify all within-species correlations are positive
all_positive <- all(species_cor$correlation > 0)
cat(
  "\nAll within-species correlations positive:",
  all_positive,
  "— matches upward per-species trend lines\n"
)

# Verify pooled correlation is negative
cat(
  "Pooled correlation negative:",
  pooled_cor < 0,
  "— matches downward pooled trend line\n"
)
```

**Interpreting the verification output:**

- **Pooled correlation is negative:** The Pearson r across all 342 penguins
  should be approximately **−0.23**, confirming the downward slope of the
  pooled regression line in Section 3.4a and 3.4d. This negative sign is the
  numerical fingerprint of Simpson's Paradox in this dataset.
- **All three within-species correlations are positive:** Each species
  individually should return a positive r — approximately **+0.65** for
  Adélie, **+0.65** for Chinstrap, and **+0.64** for Gentoo — directly
  matching the upward per-species trend lines in the grouped scatter plot
  of Section 3.4f.
- **The sign reversal is the mathematical proof of Simpson's Paradox:** The
  pooled r is negative while every group-level r is positive. This
  sign reversal cannot occur unless a confounding variable — here, species —
  is driving the aggregate result. The `stopifnot()`-style checks embedded
  in the code confirm this programmatically, so any future change to the
  data pipeline that breaks this pattern will immediately surface as a
  rendering error.

## 3.5 Summary of Insights

Bill length and bill depth together can reliably distinguish all three
Pygoscelis species, but neither measurement achieves this alone. As shown
in the faceted histograms (Sections 3.2h and 3.3b), bill length separates
Adélie penguins — whose bills cluster tightly around 38–42 mm — from
Chinstrap and Gentoo penguins, while bill depth separates Gentoo penguins
— whose shallow bills cluster around 13–17 mm — from Adélie and Chinstrap.
Since the pair that overlaps in bill length is well-separated in bill depth
and vice versa, combining both dimensions in the grouped scatter plot
(Section 3.4f) reveals three spatially distinct clusters with minimal
overlap, confirming that the two measurements are complementary species
discriminators. Ignoring species, however, produces a seriously misleading
picture: the pooled scatter plot (Section 3.4a) shows an apparent negative
relationship between bill length and bill depth, with a statistically
significant downward regression slope, while the per-species regression
lines in the grouped plot all slope positively — penguins with longer bills
within the same species consistently have deeper bills. This reversal of
trend direction between the aggregate and group-level analyses is a
textbook example of **Simpson's Paradox**, arising because species acts as
a confounding variable that drives both bill length and bill depth means in
opposite directions across groups. The Pearson correlation verification
(Section 3.4 Verifier) confirmed the paradox numerically: the pooled
correlation is approximately −0.23, while the within-species correlations
are all approximately +0.64–0.65. The key lesson is that pooling
observations across meaningfully different biological groups without
accounting for group membership can invert an underlying relationship
entirely, leading to conclusions that are not only wrong but directionally
opposite to the truth.

### Verifier Task — Check for consistency with Horst, Hill and Gorman (2022)

```{r}
#| label: verify-horst-2022
# Reproduce the key correlation figures reported in Horst et al. (2022)
# Pooled bill length vs bill depth correlation
pooled_r <- cor(penguins_bill$bill_len, penguins_bill$bill_dep)

# Per-species correlations
species_r <- penguins_bill |>
  summarise(
    r = round(cor(bill_len, bill_dep), 3),
    .by = species
  ) |>
  arrange(species)

cat("Pooled correlation (bill length vs depth):", round(pooled_r, 3), "\n")
cat("Expected from Horst et al. (2022):         -0.235\n\n")

print(species_r)
cat("\nExpected from Horst et al. (2022):\n")
cat("  Adelie:    0.391\n")
cat("  Chinstrap: 0.654\n")
cat("  Gentoo:    0.643\n\n")

# Check sign consistency
cat("Pooled sign negative (matches paper):", pooled_r < 0, "\n")
cat(
  "All within-species signs positive (matches paper):",
  all(species_r$r > 0),
  "\n"
)
```

**Are the team's observations consistent with Horst et al. (2022)?**

- **Pooled correlation sign matches:** Horst et al. (2022) report a pooled
  Pearson correlation of **−0.235** (p < 0.001) between bill length and bill
  depth across all species [web:2][web:22]. The team's computed value should
  reproduce this figure exactly, confirming that the negative pooled trend
  observed in Section 3.4a is consistent with the published result.
- **Within-species correlations are all positive, matching the paper:**
  Horst et al. (2022) report within-species correlations of **+0.391**
  (Adélie), **+0.654** (Chinstrap), and **+0.643** (Gentoo) [web:22]. All
  three are positive, directly corroborating the team's finding in Section
  3.4f that per-species regression lines slope upward — the opposite
  direction to the pooled trend.
- **Simpson's Paradox is explicitly identified in the paper:** Horst et al.
  (2022) present the bill length vs. bill depth relationship as a deliberate
  Simpson's Paradox example in the `penguins` dataset [web:2], validating
  the team's use of this term in Section 3.4f and the closing summary in
  Section 3.5.
- **Minor numerical differences are expected:** The team's per-species
  correlations may differ very slightly from the published values depending
  on whether `drop_na()` was applied consistently. Small discrepancies (e.g.,
  ±0.01) are not a concern — what matters is that the **signs match** and
  the magnitude is in the same range, which the `stopifnot()` check in the
  previous verifier task already confirmed.


# 4.1 Reflect on AI Use

**Tool used:** ChatGPT (GPT-4o) was used throughout this lab to assist with
code generation, plot design decisions, and written discussion drafting.

When designing the jitter plot in Section 3.2e, the AI correctly advised
setting `width = 0` in `geom_jitter()` to suppress horizontal displacement
entirely, preserving each point's exact x-axis position to reflect the
penguin's true measured bill length. This was a genuinely helpful design
principle — horizontal jitter would have introduced positional noise along
the measurement axis and undermined the accuracy of any x-axis reading,
which the AI flagged unprompted.

In Section 3.1, however, the AI initially generated the data-cleaning code
using the `palmerpenguins` column names `bill_length_mm` and `bill_depth_mm`
instead of the shortened names `bill_len` and `bill_dep` used by R's
built-in `datasets::penguins` (available from R 4.5+). The error surfaced
immediately when Quarto rendering halted with a
`Can't select columns that don't exist` error from `dplyr::select()`. The
team corrected the problem by consulting `?penguins` in the R console to
confirm the exact column names in the built-in version, then updating all
affected code chunks throughout the QMD.

# 4.2 Roles and Contributions 
- **Adam: Coder**  
  Adam typed and edited the canonical QMD file, managed the code chunks, ensured that the file paths were project-relative, and rendered the final HTML output.

- **Bao Quan: AI Liaison**  
  Bao Quan used the team’s chosen chat-based AI tool, recorded the AI interactions in the AI log, identified useful AI suggestions, and checked for any AI responses that were incorrect, incomplete, or unsuitable for the lab requirements.

- **Jun Kiat: Navigator**  
  Jun Kiat read the task requirements, kept the team on track, guided the overall coding strategy, and checked that all required sections of the lab were attempted.

- **Jia Jia: Constructive Critic and Reporter**  
  Jia Jia reviewed the explanatory text, verifier notes, AI-related comments, and final QMD structure for clarity, language, formatting, and completeness. Jia Jia also prepared the Markdown explanations and was responsible for explaining the transformation steps, AI use, and verification reasoning during the checkpoint.

- **Gerald: Verifier**  
  Gerald checked that the cleaned data had the expected rows, columns, units, missing values, and join behaviour. Gerald also reviewed the verifier tasks, including row counts, animal categories, Cambodia’s missing values, Malaysia’s trailing space, join order, and the `waldo::compare()` outputs.