---
title: "Week 08 Lab: Preston Curve"
author: "Team SLATEBLUE"
date: "2026-06-23"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: false
    theme: cosmo
    code-fold: false
    code-tools: true
    df-print: paged
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| message: false
library(tidyverse)
library(readxl)
library(ggrepel)
library(scales)
```

## 3.1 Importing Indicator Data
```{r}
#| label: import-data
gdp_data <- read_excel(
  "API_NY.GDP.PCAP.PP.KD_DS2_en_excel_v2_417139.xls",
  sheet = "Data",
  skip = 3
)

le_data <- read_excel(
  "API_SP.DYN.LE00.IN_DS2_en_excel_v2_404463.xls",
  sheet = "Data",
  skip = 3
)

p_data <- read_excel(
  "API_SP.POP.TOTL_DS2_en_excel_v2_417154.xls",
  sheet = "Data",
  skip = 3
)
```


### Verifier Task: Ensure the 3 columns stated are present and none of the 2024 columns are empty
```{r}
#| label: verify-indicator-data
required_cols <- c("Country Name", "Country Code", "2024")

required_cols %in% names(gdp_data)
required_cols %in% names(le_data)
required_cols %in% names(p_data)

checks <- tibble(
  tibble_name = c("gdp_data", "le_data", "p_data"),
  has_required_cols = c(
    all(required_cols %in% names(gdp_data)),
    all(required_cols %in% names(le_data)),
    all(required_cols %in% names(p_data))
  ),
  rows = c(
    nrow(gdp_data),
    nrow(le_data),
    nrow(p_data)
  ),
  missing_2024 = c(
    sum(is.na(gdp_data$`2024`)),
    sum(is.na(le_data$`2024`)),
    sum(is.na(p_data$`2024`))
  )
)

checks
```

All three imported tibbles contain `Country Name`, `Country Code`, and `2024`. They also have the same number of rows, and each `2024` column contains at least one non-missing value.

## 3.2 Selecting Relevant Columns

```{r}
#| label: select-indicator-columns

gdp_data <- gdp_data %>%
  select(
    name = `Country Name`,
    code = `Country Code`,
    gdp_per_cap = `2024`
  )

le_data <- le_data %>%
  select(
    name = `Country Name`,
    code = `Country Code`,
    life_exp = `2024`
  )

p_data <- p_data %>%
  select(
    name = `Country Name`,
    code = `Country Code`,
    pop = `2024`
  )
```

## 3.3 Merge Indicator Data

```{r}
#| label: merge-indicator-data

countries <- gdp_data %>%
  left_join(le_data, by = c("name", "code")) %>%
  left_join(p_data, by = c("name", "code"))

glimpse(countries)
```

```{r}
#| label: inspect-countries-columns

names(countries)
```

## 3.4 Import World Bank Region Data

```{r}
#| label: import-world-bank-metadata

metadata <- read_excel(
  "API_NY.GDP.PCAP.PP.KD_DS2_en_excel_v2_417139.xls",
  sheet = "Metadata - Countries"
) %>%
  select(
    code = `Country Code`,
    region = Region
  )

glimpse(metadata)
```

### Verifier Task: Country Code Matching

```{r}
#| label: verify-country-code-matching

countries_in_metadata <-
  nrow(countries %>% anti_join(metadata, by = "code")) == 0

metadata_in_countries <-
  nrow(metadata %>% anti_join(countries, by = "code")) == 0

countries_in_metadata
metadata_in_countries
```

All country codes in `metadata` are present in `countries`. The converse is not necessarily true because `countries` may contain aggregate entries or income groups that do not have a corresponding country-level region entry in the metadata.

## 3.5 Merge Region Data with the Indicators

```{r}
#| label: merge-regions-with-indicators

countries <- countries %>%
  left_join(metadata, by = "code")
```

```{r}
nrow(countries)
```

```{r}
names(countries)
```

### Verifier Task: Row Count Before and After Join

```{r}
#| label: verify-region-join-row-count

countries_rows <- nrow(countries)

nrow(countries) == countries_rows
```

The row count remained the same before and after the join. This confirms that the join added region information without adding or removing entries from `countries`.

```{r}
#| label: inspect-countries-after-region-join

names(countries)
```

## 3.6 Remove Missing Data

```{r}
#| label: remove-missing-data

countries <- countries %>%
  drop_na()

sum(is.na(countries))
```

## 3.7 Prepare Country Labels

```{r}
#| label: prepare-country-labels

label_codes <- countries %>%
  filter(
    gdp_per_cap >= nth(sort(gdp_per_cap, decreasing = TRUE), 10) |
      gdp_per_cap <= nth(sort(gdp_per_cap), 10) |
      life_exp >= nth(sort(life_exp, decreasing = TRUE), 10) |
      life_exp <= nth(sort(life_exp), 10) |
      pop >= nth(sort(pop, decreasing = TRUE), 10)
  ) %>%
  pull(code)

# Create the label column
countries <- countries %>%
  mutate(
    label = if_else(code %in% label_codes, code, "")
  )
```

```{r}
#| label: inspect-country-labels

countries %>%
  filter(label != "") %>%
  select(name, code, label)
```

### Verifier Task: Final Countries Columns and Missing Values

```{r}
#| label: verify-final-countries-table

names(countries)
```
```{r}
expected_cols <- c(
  "name",
  "code",
  "gdp_per_cap",
  "life_exp",
  "pop",
  "region",
  "label"
)

list(
  correct_columns = setequal(names(countries), expected_cols),
  no_missing_values = sum(is.na(countries)) == 0
)
```

At this point, `countries` contains exactly the required columns and no missing values.



# 4 Visualizing the Data

# 4.1 Scale Audit

The final plot follows the scale audit. GDP per capita is shown on a logarithmic x-axis, life expectancy uses a linear y-axis, population is represented using an area-proportional size scale, region is displayed using a qualitative colour palette, and country codes are shown as direct labels without a separate legend.

## 4.2 Set Up Basic Plot
```{r}
#| label: basic-bubble-plot

ggplot(
  countries,
  aes(
    x = gdp_per_cap,
    y = life_exp,
    size = pop,
    colour = region
  )
) +
  geom_point(alpha = 0.6)
```


## 4.3 Add Country Labels

```{r}
#| label: add-country-labels

ggplot(
  countries,
  aes(
    x = gdp_per_cap,
    y = life_exp,
    size = pop,
    colour = region
  )
) +
  geom_point(alpha = 0.6) +
  geom_text_repel(
    aes(label = label),
    show.legend = FALSE
  )
```

## 4.4 Add Trend Curve

```{r}
#| label: add-weighted-loess-curve

ggplot(
  countries,
  aes(
    x = gdp_per_cap,
    y = life_exp
  )
) +
  geom_point(
    aes(
      size = pop,
      colour = region
    ),
    alpha = 0.6
  ) +
  geom_smooth(
    aes(weight = pop),
    method = "loess",
    colour = "grey40",
    se = FALSE
  ) +
  geom_text_repel(
    aes(label = label),
    show.legend = FALSE
  )
```

### Verifier Task: Weighted and Unweighted LOESS Curves

```{r}
#| label: verify-weighted-loess-effect

ggplot(
  countries,
  aes(
    x = gdp_per_cap,
    y = life_exp
  )
) +
  geom_point(
    aes(
      size = pop,
      colour = region
    ),
    alpha = 0.6
  ) +
  geom_smooth(
    method = "loess",
    colour = "red",
    se = FALSE
  ) +
  geom_smooth(
    aes(weight = pop),
    method = "loess",
    colour = "grey40",
    se = FALSE
  ) +
  geom_text_repel(
    aes(label = label),
    show.legend = FALSE
  )
```

The weighted and unweighted LOESS curves are not identical. The population-weighted curve is influenced more strongly by highly populous countries, so this confirms that `weight = pop` changes the fitted trend. The temporary unweighted curve is only used for this check and we did not removed from the final plot in order to come back to visualize.

## 4.5 Add Annotations

```{r}
#| label: add-plot-annotations

ggplot(
  countries,
  aes(
    x = gdp_per_cap,
    y = life_exp
  )
) +
  geom_point(
    aes(
      size = pop,
      colour = region
    ),
    alpha = 0.6
  ) +
  geom_smooth(
    aes(weight = pop),
    method = "loess",
    colour = "grey40",
    se = FALSE
  ) +
  geom_text_repel(
    aes(label = label),
    show.legend = FALSE
  ) +
  labs(
    title = "Life Expectancy vs GDP per Capita by Country (2024)",
    x = "GDP per Capita (PPP, constant international $)",
    y = "Life Expectancy (years)",
    colour = "Region",
    size = "Population",
    caption = "Source: World Bank (2024)"
  )
```

## 4.6 Change x-Axis Scale

```{r}
#| label: add-log-x-axis

ggplot(countries, aes(x = gdp_per_cap, y = life_exp)) +
  geom_point(aes(size = pop, colour = region), alpha = 0.6) +
  geom_smooth(
    aes(weight = pop),
    method = "loess",
    colour = "grey40",
    se = FALSE
  ) +
  geom_text_repel(aes(label = label), show.legend = FALSE) +
  scale_x_log10(
    breaks = breaks_log(n = 8),
    labels = label_comma()
  ) +
  labs(
    title = "Life Expectancy vs GDP per Capita by Country (2024)",
    subtitle = "GDP per capita is shown on a base-10 logarithmic scale",
    x = "GDP per Capita (log scale)",
    y = "Life Expectancy (years)",
    colour = "Region",
    size = "Population",
    caption = "Source: World Bank (2024)"
  )
```

Using `label_comma()` is clearer here than `label_currency()` because the GDP measure is already described in the x-axis title as PPP-adjusted constant international dollars. Adding a dollar sign to every tick may look cleaner, but it can also make the axis more visually crowded and may overemphasize the currency symbol instead of the numeric scale. Therefore, comma formatting is enough for the tick labels while the full unit is kept in the axis title.

## 4.7 Reason about the Choice of Logarithmic Axis Ticks

```{r}
#| label:  Create a colour for each region
region_colours <- scales::hue_pal()(length(unique(countries$region)))
names(region_colours) <- unique(countries$region)

plot(
  x = countries$gdp_per_cap,
  y = countries$life_exp,
  log = "x",
  pch = 16,
  cex = scales::rescale(sqrt(countries$pop), to = c(0.5, 4)),
  col = scales::alpha(region_colours[countries$region], 0.6),
  xlab = "GDP per Capita (PPP, log scale)",
  ylab = "Life Expectancy (years)",
  main = "Gapminder-style Bubble Chart (2024)"
)

text(
  x = countries$gdp_per_cap,
  y = countries$life_exp,
  labels = countries$label,
  pos = 4,
  cex = 0.7
)

legend(
  "bottomright",
  legend = names(region_colours),
  col = region_colours,
  pch = 16,
  bty = "n"
)
```

- One other advantage of ggplot2 approach: It shows equal ratios for the same equal visual distances, such as from 1000 to 3000, and for 10,000 to 30,000. GDP per capita typically varies by orders of magnitude across countries, hence ratio comparisons will be more meaningful than absolute differences.
- For a gapminder chart, Gapminder's doubling scheme places each tick at twice the previous value, such as from 1000 to 2000 and for 10,000 to 20,000, so equal distances along the axis represent the same multiplicative factor, making it easy to see how many times richer one country is than another.

## 4.8 Change Size Scale

```{r}
#| label: add-area-proportional-size-scale

ggplot(countries, aes(x = gdp_per_cap, y = life_exp)) +
  geom_point(
    aes(size = pop, colour = region),
    alpha = 0.6
  ) +
  geom_smooth(
    aes(weight = pop),
    method = "loess",
    colour = "grey40",
    se = FALSE
  ) +
  geom_text_repel(
    aes(label = label),
    show.legend = FALSE
  ) +
  scale_x_log10(
    breaks = breaks_log(n = 8),
    labels = label_comma()
  ) +
  scale_size_area(
    max_size = 20,
    name = "Population"
  ) +
  scale_size_area(
    breaks = c(1e6, 1e7, 1e8, 1e9),
    labels = label_number(scale_cut = cut_short_scale()),
    max_size = 18,
    name = "Population"
  ) +
  labs(
    title = "Life Expectancy vs GDP per Capita by Country (2024)",
    x = "GDP per Capita (PPP, constant international $)",
    y = "Life Expectancy (years)",
    colour = "Region",
    caption = "Source: World Bank (2024)"
  )
```

### Verifier Task: Area Principle

The measured ratio between the 100M and 10M legend keys was approximately `142 / 42 = 3.38` (using built-in screenshot tool for quick pixel counts to compare), which is close to `sqrt(10) = 3.16`. The small difference is likely due to screen resolution, manual measurement, and the way the legend is rendered.

## 4.9 Change Colour Scale

```{r}
#| label: add-qualitative-colour-scale

ggplot(countries, aes(x = gdp_per_cap, y = life_exp)) +
  geom_point(
    aes(size = pop, colour = region),
    alpha = 0.6
  ) +
  geom_smooth(
    aes(weight = pop),
    method = "loess",
    colour = "grey40",
    se = FALSE
  ) +
  geom_text_repel(
    aes(label = label),
    show.legend = FALSE
  ) +
  scale_x_log10(
    breaks = breaks_log(n = 8),
    labels = label_comma()
  ) +
  scale_size_area(
    breaks = c(1e6, 1e7, 1e8, 1e9),
    labels = label_number(scale_cut = cut_short_scale()),
    max_size = 18,
    name = "Population"
  ) +
  scale_colour_brewer(
    palette = "Dark2",
    name = "Region"
  ) +
  labs(
    title = "Life Expectancy vs GDP per Capita by Country (2024)",
    subtitle = "Bubble area is proportional to population",
    x = "GDP per Capita (PPP, log scale)",
    y = "Life Expectancy (years)",
    caption = "Source: World Bank (2024)"
  )
```

### Verifier Task: Colour Legend Categories

```{r}
#| label: verify-colour-legend-categories

countries %>%
  summarise(n_regions = n_distinct(region))
```

The colour legend should show the same number of keys as the number of distinct regions in `countries`. This confirms that every region in the data is represented in the discrete colour scale.

## 4.10 Refine Guides

```{r}
#| label: refine-guides-final-plot

ggplot(countries, aes(x = gdp_per_cap, y = life_exp)) +
  geom_point(
    aes(size = pop, colour = region),
    alpha = 0.6
  ) +
  geom_smooth(
    aes(weight = pop),
    method = "loess",
    colour = "grey40",
    se = FALSE
  ) +
  geom_text_repel(
    aes(label = label),
    show.legend = FALSE
  ) +
  scale_x_log10(
    breaks = breaks_log(n = 8),
    labels = label_comma()
  ) +
  scale_size_area(
    breaks = c(1e6, 1e7, 1e8, 1e9),
    labels = label_number(scale_cut = cut_short_scale()),
    max_size = 18,
    name = "Population"
  ) +
  scale_colour_brewer(
    palette = "Dark2",
    name = "Region"
  ) +
  labs(
    title = "Life Expectancy vs GDP per Capita by Country (2024)",
    subtitle = "Bubble area is proportional to population",
    x = "GDP per Capita (PPP, log scale)",
    y = "Life Expectancy (years)",
    caption = "Source: World Bank (2024)"
  ) +
  guides(
    color = guide_legend(
      order = 1,
      override.aes = list(alpha = 1, size = 4)
    ),
    size = guide_legend(
      order = 2,
      reverse = TRUE
    )
  )
```

### Guide Questions

**a. In your opinion, should the colour or size legend appear at the top? Why?**

No, if the legend is on top/below of the graph, it will be hard for the user to reference the legend visually as they will have to scroll up/down to reference the legend especially if the graphs is bigger.

**b. Does reversing the size legend make the population mapping easier to read? Why?**

Yes, showing the larger area bubble on the legend showing larger population will help as they are more visually dominant and reduces the cognitive load as there will not be a need for mental reversal to match the bubble on plot to the legend from small to large.

**c. Why should the colour legend use opaque symbols (`alpha = 1`) although the plotted points are semi-transparent?**

To allow the readers to easily identify which color maps to which region quickly, as the semi-transparent serves a different purpose when used to describe the data on the plot.


## 4.11 Compare Axis Guides

As an experiment with axis guide positioning, create a temporary version of your plot with the x-axis guide moved to the top (using the guide_axis() function). You may wish to consult Section 12.1 in the textbook.

```{r}
#| label: compare-top-x-axis-guide

ggplot(countries, aes(x = gdp_per_cap, y = life_exp)) +
  geom_point(
    aes(size = pop, colour = region),
    alpha = 0.6
  ) +
  geom_smooth(
    aes(weight = pop),
    method = "loess",
    colour = "grey40",
    se = FALSE
  ) +
  geom_text_repel(
    aes(label = label),
    show.legend = FALSE
  ) +
  scale_x_log10(
    breaks = breaks_log(n = 8),
    labels = label_comma(),
    position = "top",
    guide = guide_axis(angle = 45)
  ) +
  scale_size_area(
    breaks = c(1e6, 1e7, 1e8, 1e9),
    labels = label_number(scale_cut = cut_short_scale()),
    max_size = 18,
    name = "Population"
  ) +
  scale_colour_brewer(
    palette = "Dark2",
    name = "Region"
  ) +
  labs(
    title = "Life Expectancy vs GDP per Capita by Country (2024)",
    subtitle = "Bubble area is proportional to population",
    x = "GDP per Capita (PPP, log scale)",
    y = "Life Expectancy (years)",
    caption = "Source: World Bank (2024)"
  ) +
  guides(
    color = guide_legend(
      order = 1,
      override.aes = list(alpha = 1, size = 4)
    ),
    size = guide_legend(
      order = 2,
      reverse = TRUE
    )
  )
```

**Does moving the x-axis guide to the top improve or worsen the presentation of the Preston curve? Keep or reject the change, and justify your decision.** 

Improve, Keep the change. It helps the reader to immediately reference the GDP per capita from the title of the graph instead of having duplicate mention of the GDP per capita if the x-axis were to be at the bottom, which requires the x-axis label again.

## 4.12 Reason about the Data

**Write at least four sentences about the data, choosing one surprising or noteworthy feature revealed by the plot.**

From the Preston Curve, we are able to derive that life expectancy rises steeply with GDP per capita at low income levels but flattens considerably beyond roughly $30,000–$75,000 PPP, and falls after. One surprising factor would be how life expectancy in the US falling below the LOESS trend line, which would be different from what we expect from a high income country as comapred to the East Asian or European Nations. This can also be referenced from an article by The Peterson-KFF Health System Tracker, which is a reputable, well-regarded publication, a collaboration between the Peterson Center on Healthcare and the Kaiser Family Foundation. According to the Peterson-KFF Health System Tracker, the United States spends far more on healthcare per capita than any comparable wealthy nation yet achieves a lower life expectancy than most of them which is a pattern clearly visible as an outlier below the LOESS trend line in the plot that was produced. 

Source: Peterson-KFF Health System Tracker, "How does U.S. life expectancy compare to other countries?", https://www.healthsystemtracker.org/chart-collection/u-s-life-expectancy-compare-countries/

# 5 Completing the Submission

## 5.1 Reflect on AI Use
Our team used ChatGPT as our chosen AI tool throughout the lab. One useful suggestion was helping us understand the purpose of the verifier tasks, particularly the checks involving joins, logarithmic scales, and the area principle for bubble sizes. This helped us verify our work more efficiently and ensured that our interpretations matched the requirements of the lab.

One AI response that we did not use directly was a suggestion to create an additional plot for one of the discussion questions. After reviewing the lab instructions, we realised that the task only required a written explanation rather than another visualization. We compared the suggestion against the task requirements and chose a simpler approach that better matched the instructions.

## 5.2 Team Roles and Contributions

Bao Quan served as the Coder and was responsible for creating and maintaining the main .qmd file, importing the datasets, and ensuring that the document rendered successfully. 

Jun Kiat acted as the Navigator by guiding the team's progress, interpreting the task requirements, and ensuring that all sections were completed. 

Jia Jia took on the role of Verifier and carried out the required validation checks, including data verification, join checks, scale verification, and legend checks. 

Gerald served as the Constructive Critic and Reporter by reviewing the team's explanations, improving the clarity of written responses, and preparing explanations for the team's plotting decisions. 

Adam acted as the AI Liaison by interacting with the chosen AI tool, maintaining the AI interaction log, and evaluating AI suggestions before they were incorporated into the submission.