---
title: "Week 12 Lab: Raster Maps of Sea Surface Temperature Anomalies"
author: "Team SLATEBLUE"
date: "2026-07-21"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: false
    theme: cosmo
    code-fold: false
    code-tools: true
    df-print: paged
execute:
  echo: true
  warning: false
  message: false
---

## 3.2 Start Coding
```{r}
#| label: setup
#| message: false
library(dplyr)
library(forcats)
library(ggplot2)
library(lubridate)
library(purrr)
library(readr)
library(stringr)
library(tibble)
library(tidyverse)
library(stars)
library(scales)
library(ggnewscale)
library(units)
```

## 3.3 Import the Daily Grids as stars Data Cubes
```{r}
#| label: locate-daily-files

files_2009 <- list.files(
  "oisst_lab_data/oisst_daily/el_nino_2009-12",
  pattern = "\\.nc$",
  full.names = TRUE
) |>
  sort()

files_2010 <- list.files(
  "oisst_lab_data/oisst_daily/la_nina_2010-12",
  pattern = "\\.nc$",
  full.names = TRUE
) |>
  sort()

length(files_2009)

stopifnot(
  length(files_2009) == 31,
  length(files_2010) == 31
)

c(
  first_2009 = basename(first(files_2009)),
  last_2009 = basename(last(files_2009)),
  first_2010 = basename(first(files_2010)),
  last_2010 = basename(last(files_2010))
)
```

### 3.3.1 Inspect Imported Data via read_star
```{r}
# December 2009: 31 daily grids
dec_2009 <- read_stars(files_2009, sub = c("anom", "ice"), along="time")

# December 2010: 31 daily grids
dec_2010 <- read_stars(files_2010, sub = c("anom", "ice"), along="time")

# Inspect
dec_2009
dec_2010
```

## 3.4 Establish the Coordinate Reference System
``` {r}
# Assign EPSG:4326 to both stars objects
dec_2009 <- st_set_crs(dec_2009, 4326)
dec_2010 <- st_set_crs(dec_2010, 4326)

# Check CRS
st_crs(dec_2009)
st_crs(dec_2010)

# check boundary
st_bbox(dec_2009)
st_bbox(dec_2010)
```

## 3.5 Calculate monthly summaries
## Monthly cell summaries

```{r}
#| label: monthly-summary-function

summarise_month <- function(daily_cube) {
  mean_anom <- st_apply(
    daily_cube["anom"],
    c("x", "y"),
    mean,
    na.rm = TRUE
  )
  names(mean_anom) <- "mean_anom"

  ice_zero <- daily_cube["ice"]
  ice_zero[[1]][is.na(ice_zero[[1]])] <- 0

  mean_ice <- st_apply(
    ice_zero,
    c("x", "y"),
    mean
  )
  names(mean_ice) <- "mean_ice"

  result <- c(mean_anom, mean_ice) |>
    mutate(
      land = is.nan(mean_anom) & mean_ice == 0,
      mean_ice = replace(mean_ice, land, NA_real_)
    )

  result[c("land", "mean_anom", "mean_ice")]
}
```

```{r}
#| label: calculate-monthly-summaries

summary_2009 <- summarise_month(dec_2009)
summary_2010 <- summarise_month(dec_2010)
```

## Verifier Task
```{r}
#| label: verify-monthly-summaries

verify_monthly_summary <- function(monthly, daily) {
  non_land_ice <- monthly$mean_ice[!monthly$land]

  stopifnot(
    all(dim(monthly) == c(1440, 720)),
    identical(names(st_dimensions(monthly)), c("x", "y")),
    identical(names(monthly), c("land", "mean_anom", "mean_ice")),
    !anyNA(monthly$land),
    all(is.na(monthly$mean_ice) == monthly$land),
    min(non_land_ice, na.rm = TRUE) >= 0,
    max(non_land_ice, na.rm = TRUE) <= 1,
    isTRUE(all.equal(st_bbox(monthly), st_bbox(daily))),
    isTRUE(all.equal(st_res(monthly), st_res(daily))),
    st_crs(monthly) == st_crs(daily)
  )

  tibble(
    dimensions = paste(dim(monthly), collapse = " × "),
    land_cells = sum(monthly$land),
    non_land_cells = sum(!monthly$land),
    minimum_ice = min(non_land_ice, na.rm = TRUE),
    maximum_ice = max(non_land_ice, na.rm = TRUE)
  )
}

bind_rows(
  `2009` = verify_monthly_summary(summary_2009, dec_2009),
  `2010` = verify_monthly_summary(summary_2010, dec_2010),
  .id = "year"
)
```

## 3.6 Derive Surface Classes

```{r}
# Derive Surface Classes

add_surface_class <- function(monthly) {
  # start with everything as "open ocean"
  surface_values <- rep("open ocean", length(monthly$land))
  
  # land first
  surface_values[monthly$land] <- "land"
  
  # sea ice: mean_ice >= 0.15 OR anomaly missing
  surface_values[!monthly$land & 
                 (monthly$mean_ice >= 0.15 | is.na(monthly$mean_anom))] <- "sea ice"
  
  # turn into factor with correct levels
  surface_values <- factor(surface_values,
                           levels = c("land", "sea ice", "open ocean"))
  
  # reshape back to the same dimensions as land
  dim(surface_values) <- dim(monthly$land)
  
  # add as new attribute
  monthly$surface <- surface_values
  monthly
}

summary_2009 <- add_surface_class(summary_2009)
summary_2010 <- add_surface_class(summary_2010)
```

### 3.6 Verifier Task
``` {r}
## 1. All surface values are valid and non-missing
all(summary_2009$surface %in% c("land", "sea ice", "open ocean"))
all(summary_2010$surface %in% c("land", "sea ice", "open ocean"))

## 2. Every cell with mean_ice >= 0.15 is "sea ice",
##    and every non-land cell with missing mean_anom is "sea ice"
all(summary_2009$surface[summary_2009$mean_ice >= 0.15] == "sea ice")
all(summary_2010$surface[summary_2010$mean_ice >= 0.15] == "sea ice")

all(summary_2009$surface[!summary_2009$land & is.na(summary_2009$mean_anom)] == "sea ice")
all(summary_2010$surface[!summary_2010$land & is.na(summary_2010$mean_anom)] == "sea ice")

## 3. Every "open ocean" cell has nonmissing mean_anom and mean_ice < 0.15
all(!is.na(summary_2009$mean_anom[summary_2009$surface == "open ocean"]))
all(!is.na(summary_2010$mean_anom[summary_2010$surface == "open ocean"]))

all(summary_2009$mean_ice[summary_2009$surface == "open ocean"] < 0.15)
all(summary_2010$mean_ice[summary_2010$surface == "open ocean"] < 0.15)
```

## 3.7 Combine the Monthly Summaries Along a Year Dimension
``` {r}
# Combine along new dimension
combined <- c(summary_2009[c("mean_anom", "surface")],
              summary_2010[c("mean_anom", "surface")],
              along = 3)

# Rename the third dimension to "year" and assign values
combined <- st_set_dimensions(
  combined,
  which = 3,                # the third dimension
  names = "year",           # new dimension name
  values = c(2009, 2010)    # values for that dimension
)

# Inspect
st_dimensions(combined)
```

## 3.7 Verifer
``` {r}

# 1. Dimensions are named x, y, and year with lengths 1440, 720, and 2
st_dimensions(combined)

# 2. Year values are 2009 and 2010
st_get_dimension_values(combined, "year")
# Expected: 2009 2010

# 3. Attributes are exactly mean_anom and surface
names(combined)
# Expected: "mean_anom" "surface"

# 4. Bounding box, resolution, and CRS match the originals
st_bbox(combined)
st_dimensions(combined)$x$delta
st_dimensions(combined)$y$delta
st_crs(combined)

```

## 3.8 Draft a First Faceted Raster Map
```{r}
#| label: draft-faceted-map
#| fig-width: 10
#| fig-height: 8
dec_summaries <- combined

ggplot() +
  geom_stars(data = dec_summaries, aes(fill = mean_anom)) +
  facet_wrap(~year, ncol = 1)
```

## 3.9 Design a Diverging Fill Scale

```{r}
#| label: create-open-ocean-mask

open_ocean_anom <- dec_summaries$mean_anom

open_ocean_anom[
  dec_summaries$surface != "open ocean"
] <- NA_real_

dec_summaries$open_ocean_anom <- open_ocean_anom

# Flatten both years' anomalies into one vector
anom_values <- c(dec_summaries$open_ocean_anom)

# Compute 99th percentile of absolute values
limit <- quantile(abs(anom_values), probs = 0.99, na.rm = TRUE)

limit
```

### 3.9.1 Plot
```{r}
#| label: diverging-anomaly-map
#| fig-width: 10
#| fig-height: 8

ggplot() +
  geom_stars(data = dec_summaries, aes(fill = mean_anom)) +
  facet_wrap(vars(year), ncol = 1) +
  scale_fill_distiller(
    name = "SST anomaly (°C)",
    type = "div",          # diverging palette
    palette = "RdYlBu",    # ColorBrewer RdYlBu
    direction = -1,        # reverse so blue = negative, red = positive
    limits = c(-limit, limit),  # symmetric endpoints
    oob = squish,          # squish values outside limits
    na.value = "transparent"    # land/sea ice left neutral
  ) +
  theme_minimal()
```

## 3.10 Add Land and Sea Ice as Surface Layers
``` {r}
# Define mapping from year values to labels
year_labels <- c(
  "2009" = "December 2009 (El Niño)",
  "2010" = "December 2010 (La Niña)"
)

# Define surface colours
surface_colours <- c(
  "land"       = "grey70",
  "sea ice"    = "white",
  "open ocean" = "transparent"
)

surface_colours
```

``` {r}
# Only land and sea ice should appear in the legend
surface_breaks <- c("land", "sea ice")

# Custom labels for those categories
surface_labels <- c(
  "land"    = "Land",
  "sea ice" = "Sea-ice concentration ≥ 15%"
)

surface_breaks
surface_labels
```

### 3.10.1 Plot
``` {r}
ggplot() +
  geom_stars(data = dec_summaries, aes(fill = surface)) +
  scale_fill_manual(
    values = surface_colours,
    breaks = surface_breaks,
    labels = surface_labels,
    drop = FALSE,
    na.value = "transparent"
  )

```

## 3.11 Compare Equal-Area World Projections
``` {r}
# Equal Earth projection (central meridian at 180°)
equal_earth_crs <- st_crs(
  "+proj=eqearth +lon_0=180 +ellps=WGS84 +R_A +units=m +no_defs +type=crs"
)

# Mollweide projection (central meridian at 180°)
mollweide_crs <- st_crs(
  "+proj=moll +lon_0=180 +ellps=WGS84 +R_A +units=m +no_defs +type=crs"
)

equal_earth_crs
mollweide_crs
```

