---
title: "Week 12 Lab: Raster Maps of Sea Surface Temperature Anomalies"
author: "Team SLATEBLUE"
date: "2026-07-22"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: false
    theme: cosmo
    code-fold: false
    code-tools: true
    df-print: paged
execute:
  echo: true
  warning: false
  message: false
---

## 3.1 Plan the Map Encodings

### a. Why should anom, rather than sst, determine the principal fill colors?

Sea-surface temperature (sst) varies naturally with latitude and location, so a map of absolute temperatures is dominated by geographic differences. In contrast, anom shows how much the temperature differs from the normal value at each location, making unusual warming and cooling patterns such as El Niño and La Niña easier to identify.

### b. Should the anomaly scale be sequential or diverging? What does its central value represent?

A diverging scale should be used because anomalies can be both positive and negative. Its central value is 0 °C, representing no difference from the climatological average, with colours on either side showing warmer-than-normal and cooler-than-normal conditions.

### c. Why should both panels use identical anomaly-scale limits?

Both panels should use the same anomaly-scale limits so that the same colour always represents the same temperature difference. Using different limits would make the two years difficult to compare because identical anomalies could appear as different colours.


## 3.2 Start Coding
```{r}
#| label: setup
#| message: false
library(dplyr)
library(forcats)
library(ggplot2)
library(lubridate)
library(purrr)
library(readr)
library(stringr)
library(tibble)
library(tidyverse)
library(stars)
library(scales)
library(ggnewscale)
library(units)
```

## 3.3 Import the Daily Grids as stars Data Cubes
```{r}
#| label: locate-daily-files

files_2009 <- list.files(
  "oisst_daily/el_nino_2009-12",
  pattern = "\\.nc$",
  full.names = TRUE
) |>
  sort()

files_2010 <- list.files(
  "oisst_daily/la_nina_2010-12",
  pattern = "\\.nc$",
  full.names = TRUE
) |>
  sort()

length(files_2009)

stopifnot(
  length(files_2009) == 31,
  length(files_2010) == 31
)

c(
  first_2009 = basename(first(files_2009)),
  last_2009 = basename(last(files_2009)),
  first_2010 = basename(first(files_2010)),
  last_2010 = basename(last(files_2010))
)
```

### 3.3.1 Inspect Imported Data via read_star
```{r}
# December 2009: 31 daily grids
dec_2009 <- read_stars(files_2009, sub = c("anom", "ice"), along = "time")

# December 2010: 31 daily grids
dec_2010 <- read_stars(files_2010, sub = c("anom", "ice"), along = "time")
```

### Verifier Task: Imported Daily Grids
```{r}
#| label: verify-imported-daily-grids

names(dec_2009)
dim(dec_2009)
names(st_dimensions(dec_2009))
st_get_dimension_values(dec_2009, "time")

names(dec_2010)
dim(dec_2010)
names(st_dimensions(dec_2010))
st_get_dimension_values(dec_2010, "time")
```

Both `dec_2009` and `dec_2010` contain exactly the two required attributes, `anom` and `ice`. Each object has dimensions 1440 × 720 × 31, named `x`, `y`, and `time`. The time values are arranged chronologically from December 1 to December 31 for the correct year.

## 3.4 Establish the Coordinate Reference System
``` {r}
# Assign EPSG:4326 to both stars objects
dec_2009 <- st_set_crs(dec_2009, 4326)
dec_2010 <- st_set_crs(dec_2010, 4326)
```

### Verifier Task: Coordinate Reference System
```{r}
#| label: verify-crs

# Confirm the CRS has been assigned correctly
st_crs(dec_2009) == st_crs(4326)
st_crs(dec_2010) == st_crs(4326)

# Confirm the bounding boxes are unchanged
st_bbox(dec_2009)
st_bbox(dec_2010)
```

The CRS for both `dec_2009` and `dec_2010` was successfully assigned as WGS 84 (EPSG:4326), with both verification expressions returning `TRUE`. The bounding boxes remained unchanged (`xmin = 0`, `ymin = -90`, `xmax = 360`, `ymax = 90`), confirming that assigning the CRS only added spatial reference information and did not modify the coordinate values.

## 3.5 Calculate Monthly cell summaries

```{r}
#| label: monthly-summary-function

summarise_month <- function(daily_cube) {
  mean_anom <- st_apply(
    daily_cube["anom"],
    c("x", "y"),
    mean,
    na.rm = TRUE
  )
  names(mean_anom) <- "mean_anom"

  # Treat missing daily sea-ice values as zero
  ice_zero <- daily_cube["ice"]
  ice_zero[[1]][is.na(ice_zero[[1]])] <- 0

  # Calculate mean sea-ice concentration across all 31 days
  mean_ice <- st_apply(
    ice_zero,
    c("x", "y"),
    mean
  )
  names(mean_ice) <- "mean_ice"

  # Combine the summaries and identify land cells
  result <- c(mean_anom, mean_ice) |>
    mutate(
      land = is.nan(mean_anom) & mean_ice == 0,
      mean_ice = replace(mean_ice, land, NA_real_)
    )

  result[c("land", "mean_anom", "mean_ice")]
}
```

```{r}
#| label: calculate-monthly-summaries

summary_2009 <- summarise_month(dec_2009)
summary_2010 <- summarise_month(dec_2010)
```

### Verifier Task: Monthly Cell Summaries
```{r}
#| label: verify-monthly-summaries

verify_monthly_summary <- function(monthly, daily) {
  non_land_ice <- monthly$mean_ice[!monthly$land]

  stopifnot(
    all(dim(monthly) == c(1440, 720)),
    identical(names(st_dimensions(monthly)), c("x", "y")),
    identical(names(monthly), c("land", "mean_anom", "mean_ice")),
    !anyNA(monthly$land),
    all(is.na(monthly$mean_ice) == monthly$land),
    min(non_land_ice, na.rm = TRUE) >= 0,
    max(non_land_ice, na.rm = TRUE) <= 1,
    isTRUE(all.equal(st_bbox(monthly), st_bbox(daily))),
    isTRUE(all.equal(st_res(monthly), st_res(daily))),
    st_crs(monthly) == st_crs(daily)
  )

  tibble(
    dimensions = paste(dim(monthly), collapse = " × "),
    land_cells = sum(monthly$land),
    non_land_cells = sum(!monthly$land),
    minimum_ice = min(non_land_ice, na.rm = TRUE),
    maximum_ice = max(non_land_ice, na.rm = TRUE)
  )
}

bind_rows(
  `2009` = verify_monthly_summary(summary_2009, dec_2009),
  `2010` = verify_monthly_summary(summary_2010, dec_2010),
  .id = "year"
)
```

Both monthly summary objects have dimensions 1440 × 720 and no longer contain a time dimension. They contain exactly the attributes land, mean_anom, and mean_ice. The verification checks also confirm that land has no missing values, mean_ice is missing only at land cells and otherwise lies between 0 and 1, and the spatial extent, resolution, and CRS remain unchanged from the corresponding daily data cubes.

## 3.6 Derive Surface Classes

```{r}
# Derive Surface Classes

add_surface_class <- function(monthly) {
  # start with everything as "open ocean"
  surface_values <- rep("open ocean", length(monthly$land))

  # land first
  surface_values[monthly$land] <- "land"

  # sea ice: mean_ice >= 0.15 OR anomaly missing
  surface_values[
    !monthly$land &
      (monthly$mean_ice >= 0.15 | is.na(monthly$mean_anom))
  ] <- "sea ice"

  # turn into factor with correct levels
  surface_values <- factor(
    surface_values,
    levels = c("land", "sea ice", "open ocean")
  )

  # reshape back to the same dimensions as land
  dim(surface_values) <- dim(monthly$land)

  # add as new attribute
  monthly$surface <- surface_values
  monthly
}

summary_2009 <- add_surface_class(summary_2009)
summary_2010 <- add_surface_class(summary_2010)
```

### Verifier Task: Surface Classes

```{r}
## 1. All surface values are valid and non-missing
!anyNA(summary_2009$surface)
!anyNA(summary_2010$surface)

all(summary_2009$surface %in% c("land", "sea ice", "open ocean"))
all(summary_2010$surface %in% c("land", "sea ice", "open ocean"))

## 2. Every cell with mean_ice >= 0.15 is "sea ice",
##    and every non-land cell with missing mean_anom is "sea ice"
all(summary_2009$surface[summary_2009$mean_ice >= 0.15] == "sea ice")
all(summary_2010$surface[summary_2010$mean_ice >= 0.15] == "sea ice")

all(
  summary_2009$surface[
    !summary_2009$land & is.na(summary_2009$mean_anom)
  ] ==
    "sea ice"
)

all(
  summary_2010$surface[
    !summary_2010$land & is.na(summary_2010$mean_anom)
  ] ==
    "sea ice"
)

## 3. Every "open ocean" cell has non-missing mean_anom
##    and mean_ice < 0.15
!anyNA(
  summary_2009$mean_anom[
    summary_2009$surface == "open ocean"
  ]
)

!anyNA(
  summary_2010$mean_anom[
    summary_2010$surface == "open ocean"
  ]
)

all(
  summary_2009$mean_ice[
    summary_2009$surface == "open ocean"
  ] <
    0.15
)

all(
  summary_2010$mean_ice[
    summary_2010$surface == "open ocean"
  ] <
    0.15
)
```

Both monthly summary objects contain only the three valid surface classes: land, sea ice, and open ocean, with no missing classifications. All cells with a monthly mean sea-ice concentration of at least 0.15, as well as non-land cells with a missing monthly anomaly, are classified as sea ice. Every open ocean cell has a non-missing anomaly and a monthly mean sea-ice concentration below 0.15.

## 3.7 Combine the Monthly Summaries Along a Year Dimension
``` {r}
# Combine along new dimension
dec_summaries <- c(
  summary_2009[c("mean_anom", "surface")],
  summary_2010[c("mean_anom", "surface")],
  along = 3
)

# Rename the third dimension to "year" and assign values
dec_summaries <- st_set_dimensions(
  dec_summaries,
  which = 3, # the third dimension
  names = "year", # new dimension name
  values = c(2009, 2010) # values for that dimension
)

# Inspect
st_dimensions(dec_summaries)
```

### Verifier Task: Combined Monthly Summaries
``` {r}

# 1. Dimensions are named x, y, and year with lengths 1440, 720, and 2
st_dimensions(dec_summaries)
dim(dec_summaries)

# 2. Year values are 2009 and 2010
st_get_dimension_values(dec_summaries, "year")

# 3. Attributes are exactly mean_anom and surface
names(dec_summaries)

# 4. Bounding box, resolution, and CRS match the originals
isTRUE(all.equal(st_bbox(dec_summaries), st_bbox(summary_2009)))
isTRUE(all.equal(st_bbox(dec_summaries), st_bbox(summary_2010)))

isTRUE(all.equal(st_res(dec_summaries), st_res(summary_2009)))
isTRUE(all.equal(st_res(dec_summaries), st_res(summary_2010)))

st_crs(dec_summaries) == st_crs(summary_2009)
st_crs(dec_summaries) == st_crs(summary_2010)
```

The combined object, dec_summaries, has dimensions x, y, and year, with lengths 1440, 720, and 2. The year dimension contains the values 2009 and 2010, and the object contains only the mean_anom and surface attributes. Its bounding box, spatial resolution, and CRS are unchanged from both monthly summary objects.

## 3.8 Draft a First Faceted Raster Map
```{r}
#| label: draft-faceted-map
#| fig-width: 10
#| fig-height: 8

ggplot() +
  geom_stars(data = dec_summaries, aes(fill = mean_anom)) +
  facet_wrap(~year, ncol = 1)
```

Shortcoming List:
1. The default colour scale is not centred at 0 °C, making it difficult to distinguish positive anomalies from negative anomalies.
2. Land and sea ice are both displayed in grey, so they cannot be distinguished from one another.
3. The map lacks a title, subtitle and contextual information, making it difficult for readers to understand what the map represents.
4. Regions near the poles appear much larger than comparable regions near the equator, making comparisons of the spatial extent of anomalies across different latitudes misleading.


## 3.9 Design a Diverging Fill Scale

```{r}
#| label: create-open-ocean-mask

open_ocean_anom <- dec_summaries$mean_anom

open_ocean_anom[
  dec_summaries$surface != "open ocean"
] <- NA_real_

dec_summaries$open_ocean_anom <- open_ocean_anom

# Flatten both years' anomalies into one vector
anom_values <- c(dec_summaries$open_ocean_anom)

# Compute 99th percentile of absolute values
limit <- ceiling(
  quantile(abs(anom_values), probs = 0.99, na.rm = TRUE)
)

limit
```

### 3.9.1 Plot
```{r}
#| label: diverging-anomaly-map
#| fig-width: 10
#| fig-height: 8

ggplot() +
  geom_stars(data = dec_summaries, aes(fill = open_ocean_anom)) +
  facet_wrap(vars(year), ncol = 1) +
  scale_fill_distiller(
    name = "SST anomaly (°C)",
    type = "div", # diverging palette
    palette = "RdYlBu", # ColorBrewer RdYlBu
    direction = -1, # reverse so blue = negative, red = positive
    limits = c(-limit, limit), # symmetric endpoints
    oob = scales::squish, # squish values outside limits
    na.value = "transparent" # land/sea ice left neutral
  )
```

## 3.10 Add Land and Sea Ice as Surface Layers
``` {r}
#| label: define-surface-colours

# Define surface colours
surface_colours <- c(
  "land" = "grey70",
  "sea ice" = "white",
  "open ocean" = "transparent"
)

surface_colours
```

``` {r}
#| label: define-surface-legend
# Only land and sea ice should appear in the legend
surface_breaks <- c("land", "sea ice")

# Custom labels for those categories
surface_labels <- c(
  "land" = "Land",
  "sea ice" = "Sea-ice concentration ≥ 15%"
)

surface_breaks
surface_labels
```

### 3.10.1 Plot
``` {r}
#| label: plot-surface-layers
ggplot() +
  # Draw SST anomalies first
  geom_stars(
    data = dec_summaries,
    aes(fill = open_ocean_anom)
  ) +
  scale_fill_distiller(
    name = "SST anomaly (°C)",
    type = "div",
    palette = "RdYlBu",
    direction = -1,
    limits = c(-limit, limit),
    oob = scales::squish,
    na.value = "transparent"
  ) +

  # Start a new fill scale for surface types
  new_scale_fill() +

  # Draw land and sea ice above the anomalies
  geom_stars(
    data = dec_summaries,
    aes(fill = surface)
  ) +
  scale_fill_manual(
    name = "Surface",
    values = surface_colours,
    breaks = surface_breaks,
    labels = surface_labels,
    drop = FALSE,
    na.value = "transparent"
  ) +
  facet_wrap(~year, ncol = 1)
```

The surface layer must be drawn above the anomaly layer so that land and sea ice cover any anomaly values beneath them. Land is shown in grey and sea ice in white, while open-ocean cells are transparent. This allows the anomaly colours to remain visible only over open ocean.

### Verifier Task: Surface Layer Mask

```{r}
#| label: verify-surface-mask

verify_surface_mask <- function(data) {
  open_ocean <- data$surface == "open ocean"
  other_surface <- data$surface != "open ocean"

  stopifnot(
    # Open-ocean cells must keep the original anomaly
    isTRUE(all.equal(
      data$open_ocean_anom[open_ocean],
      data$mean_anom[open_ocean]
    )),

    # Non-open-ocean cells must be missing
    all(is.na(data$open_ocean_anom[other_surface])),

    # Open-ocean cells must never have a missing anomaly
    !anyNA(data$open_ocean_anom[open_ocean]),

    # Surface values must only contain the expected classes
    all(data$surface %in% c("land", "sea ice", "open ocean"))
  )

  tibble(
    open_ocean_cells = sum(open_ocean),
    land_cells = sum(data$surface == "land"),
    sea_ice_cells = sum(data$surface == "sea ice"),
    missing_open_ocean_anomalies = sum(is.na(data$open_ocean_anom[open_ocean])),
    non_ocean_values_retained = sum(!is.na(data$open_ocean_anom[other_surface]))
  )
}

verify_surface_mask(dec_summaries)
```

The verification confirms that every open-ocean cell retains the same value in open_ocean_anom as in mean_anom, while all land and sea-ice cells are set to missing. It also confirms that no open-ocean cell has a missing anomaly. This is guaranteed by the surface-class rules because a cell is classified as open ocean only when mean_anom is available and the mean sea-ice concentration is below 0.15.

## 3.11 Compare Equal-Area World Projections
``` {r}
#| label: define-projections

# Equal Earth projection (central meridian at 180°)
equal_earth_crs <- st_crs(
  "+proj=eqearth +lon_0=180 +ellps=WGS84 +R_A +units=m +no_defs +type=crs"
)

# Mollweide projection (central meridian at 180°)
mollweide_crs <- st_crs(
  "+proj=moll +lon_0=180 +ellps=WGS84 +R_A +units=m +no_defs +type=crs"
)
```

```{r}
#| label: create-world-boundary

create_world_boundary <- function(crs) {
  longitude <- seq(0.001, 359.999, by = 1)
  latitude <- seq(-89.5, 89.5, by = 1)

  bottom <- cbind(longitude, -89.5)
  right <- cbind(359.999, latitude)
  top <- cbind(rev(longitude), 89.5)
  left <- cbind(0.001, rev(latitude))

  boundary_coordinates <- rbind(
    bottom,
    right[-1, ],
    top[-1, ],
    left[-1, ],
    bottom[1, ]
  )

  st_sfc(
    st_polygon(list(boundary_coordinates)),
    crs = 4326
  ) |>
    st_transform(crs)
}
```

```{r}
#| label: warp-raster-projection

warp_projection <- function(data, crs) {
  # Keep raster rows between -89.5° and 89.5°
  y_values <- st_get_dimension_values(data, "y")
  keep <- which(abs(y_values) <= 89.5)

  trimmed <- data[,, keep, ]

  # Build the target grid from the transformed raster extent
  target <- st_as_stars(
    st_bbox(st_transform(trimmed, crs)),
    dx = 50000,
    dy = 50000
  )

  # Reproject using nearest-neighbour resampling
  warped <- st_warp(
    trimmed,
    target,
    method = "near"
  )

  warped
}
```

```{r}
#| label: create-projected-rasters

equal_earth_map <- warp_projection(
  dec_summaries,
  equal_earth_crs
)

mollweide_map <- warp_projection(
  dec_summaries,
  mollweide_crs
)
```

```{r}
#| label: plot-projection-function

plot_projection <- function(projected_map, projection_title) {
  ggplot() +
    geom_stars(
      data = projected_map,
      aes(fill = open_ocean_anom)
    ) +
    scale_fill_distiller(
      name = "SST anomaly (°C)",
      type = "div",
      palette = "RdYlBu",
      direction = -1,
      limits = c(-limit, limit),
      oob = scales::squish,
      na.value = "transparent"
    ) +

    new_scale_fill() +

    geom_stars(
      data = projected_map,
      aes(fill = surface)
    ) +
    scale_fill_manual(
      name = "Surface",
      values = surface_colours,
      breaks = surface_breaks,
      labels = surface_labels,
      drop = FALSE,
      na.value = "transparent"
    ) +

    facet_wrap(
      ~year,
      ncol = 1
    ) +
    coord_equal() +
    labs(title = projection_title) +
    theme_minimal() +
    theme(
      axis.title = element_blank()
    )
}
```

```{r}
#| label: plot-equal-earth
#| fig-width: 10
#| fig-height: 8

plot_projection(
  equal_earth_map,
  "Equal Earth projection"
)
```

```{r}
#| label: plot-mollweide
#| fig-width: 10
#| fig-height: 8

plot_projection(
  mollweide_map,
  "Mollweide projection"
)
```

### a. How does each projection represent the polar regions?
Both projections reduce the size distortion near the poles compared to the original map. The Equal Earth projection has a flatter shape near the poles, while the Mollweide projection has a more rounded outline.

### b. How clearly does each projection show the equatorial Pacific?
Both projections show the equatorial Pacific clearly because it is placed in the centre of the map. The warm anomalies in 2009 and cool anomalies in 2010 are easy to compare in both projections.

### c. Which projection uses the rectangular panel space more effectively?
The Equal Earth projection uses the available plotting space more effectively because it fills more of the rectangular panel. The Mollweide projection leaves more empty space around the corners.

### d. Which projection makes the contrast between the two Decembers more visually prominent?
Both projections show the differences between December 2009 and December 2010 clearly. However, the Mollweide projection makes the Pacific pattern slightly easier to compare because its rounded global shape gives greater visual emphasis to the central equatorial region.

### Final projection choice
Our team chose the Mollweide projection for the final map. Although it leaves more empty space around the corners, it preserves area and makes the El Niño and La Niña patterns in the Pacific Ocean easier to compare.

## 3.12 Build the Two-Panel Anomaly Map

```{r}
#| label: prepare-final-map-elements

# Create the Mollweide map boundary
mollweide_boundary <- create_world_boundary(mollweide_crs)

# Create a line following the equator
equator <- st_sfc(
  st_linestring(
    cbind(
      seq(0.001, 359.999, by = 1),
      0
    )
  ),
  crs = 4326
) |>
  st_transform(mollweide_crs)

# Labels for the two facet panels
year_labels <- c(
  "2009" = "December 2009 (El Niño)",
  "2010" = "December 2010 (La Niña)"
)
```

```{r}
#| label: final-continuous-anomaly-map
#| fig-width: 11
#| fig-height: 9

continuous_map <- ggplot() +

  # Open-ocean anomaly layer
  geom_stars(
    data = mollweide_map,
    aes(fill = open_ocean_anom)
  ) +
  scale_fill_distiller(
    name = "SST anomaly (°C)",
    type = "div",
    palette = "RdYlBu",
    direction = -1,
    limits = c(-limit, limit),
    oob = scales::squish,
    na.value = "transparent",
    guide = guide_colourbar(
      order = 1,
      frame.colour = "grey20",
      frame.linewidth = 0.4,
      ticks.colour = "grey20"
    )
  ) +

  # Start a second fill scale
  new_scale_fill() +

  # Land and sea-ice surface layer
  geom_stars(
    data = mollweide_map,
    aes(fill = surface)
  ) +
  scale_fill_manual(
    name = "Surface",
    values = surface_colours,
    breaks = surface_breaks,
    labels = surface_labels,
    drop = FALSE,
    na.value = "transparent",
    guide = guide_legend(
      order = 2,
      override.aes = list(
        colour = "grey20",
        linewidth = 0.4
      )
    )
  ) +

  # Draw the equator and projection boundary
  geom_sf(
    data = equator,
    colour = "grey35",
    linewidth = 0.3,
    alpha = 0.6,
    inherit.aes = FALSE
  ) +
  geom_sf(
    data = mollweide_boundary,
    fill = NA,
    colour = "grey20",
    linewidth = 0.4,
    inherit.aes = FALSE
  ) +

  # Create one vertically stacked panel per year
  facet_wrap(
    ~year,
    ncol = 1,
    labeller = labeller(
      year = as_labeller(year_labels)
    )
  ) +

  # Keep the Mollweide shape and remove geographic grid labels
  coord_sf(
    crs = mollweide_crs,
    datum = NA,
    expand = FALSE
  ) +

  labs(
    title = "Sea-Surface Temperature Anomalies",
    subtitle = paste0(
      "December 2009 and 2010 compared with the ",
      "1971–2000 climatological average"
    ),
    caption = "Source: NOAA Optimum Interpolation Sea Surface Temperature (OISST)"
  ) +

  theme_void() +
  theme(
    plot.title = element_text(
      face = "bold",
      size = 17
    ),
    plot.subtitle = element_text(
      size = 11,
      margin = margin(b = 10)
    ),
    plot.caption = element_text(
      size = 9,
      colour = "grey35"
    ),
    strip.text = element_text(
      face = "bold",
      size = 11,
      margin = margin(t = 4, b = 4)
    ),
    panel.spacing = unit(0.8, "lines"),
    legend.position = "right",
    legend.box = "vertical",
    legend.title = element_text(face = "bold"),
    plot.margin = margin(10, 10, 10, 10)
  )

continuous_map
```

## 3.13 Bin the Anomaly Scale

```{r}
#| label: create-anomaly-bins

# Create whole-degree boundaries between -limit and +limit
bounded_breaks <- seq(
  from = -limit,
  to = limit,
  by = 1
)

# Add an open-ended class at each side
anomaly_breaks <- c(
  -Inf,
  bounded_breaks,
  Inf
)

# Create readable labels for the classes
bounded_labels <- paste0(
  bounded_breaks[-length(bounded_breaks)],
  " to < ",
  bounded_breaks[-1],
  " °C"
)

anomaly_labels <- c(
  paste0("< ", -limit, " °C"),
  bounded_labels,
  paste0("≥ ", limit, " °C")
)

anomaly_labels
```

```{r}
#| label: add-binned-anomaly-attribute

# Make a copy so the continuous map remains unchanged
mollweide_binned_map <- mollweide_map

# Convert the anomaly array into a vector before using cut()
binned_anom <- cut(
  as.vector(mollweide_binned_map$open_ocean_anom),
  breaks = anomaly_breaks,
  labels = anomaly_labels,
  right = FALSE,
  include.lowest = TRUE
)

# Restore the raster dimensions
dim(binned_anom) <- dim(
  mollweide_binned_map$open_ocean_anom
)

# Add the binned anomaly as a new raster attribute
mollweide_binned_map$binned_anom <- binned_anom
```

```{r}
#| label: verify-anomaly-bins

# Inspect the factor levels
levels(mollweide_binned_map$binned_anom)

# Confirm that all non-missing open-ocean anomalies received a class
all(
  !is.na(
    mollweide_binned_map$binned_anom[
      !is.na(mollweide_binned_map$open_ocean_anom)
    ]
  )
)

# Confirm that masked cells remain missing
all(
  is.na(
    mollweide_binned_map$binned_anom[
      is.na(mollweide_binned_map$open_ocean_anom)
    ]
  )
)
```

```{r}
#| label: final-binned-anomaly-map
#| fig-width: 11
#| fig-height: 9

binned_map <- ggplot() +

  # Binned open-ocean anomaly layer
  geom_stars(
    data = mollweide_binned_map,
    aes(fill = binned_anom)
  ) +
  scale_fill_brewer(
    name = "SST anomaly (°C)",
    type = "div",
    palette = "RdYlBu",
    direction = -1,
    drop = FALSE,
    na.translate = FALSE,
    na.value = "transparent",
    guide = guide_legend(
      order = 1,
      reverse = TRUE,
      override.aes = list(
        colour = "grey20",
        linewidth = 0.4
      )
    )
  ) +

  new_scale_fill() +

  # Keep the same surface layer as Section 3.12
  geom_stars(
    data = mollweide_binned_map,
    aes(fill = surface)
  ) +
  scale_fill_manual(
    name = "Surface",
    values = surface_colours,
    breaks = surface_breaks,
    labels = surface_labels,
    drop = FALSE,
    na.value = "transparent",
    guide = guide_legend(
      order = 2,
      override.aes = list(
        colour = "grey20",
        linewidth = 0.4
      )
    )
  ) +

  # Keep the same equator and boundary
  geom_sf(
    data = equator,
    colour = "grey35",
    linewidth = 0.3,
    alpha = 0.6,
    inherit.aes = FALSE
  ) +
  geom_sf(
    data = mollweide_boundary,
    fill = NA,
    colour = "grey20",
    linewidth = 0.4,
    inherit.aes = FALSE
  ) +

  facet_wrap(
    ~year,
    ncol = 1,
    labeller = labeller(
      year = as_labeller(year_labels)
    )
  ) +

  coord_sf(
    crs = mollweide_crs,
    datum = NA,
    expand = FALSE
  ) +

  labs(
    title = "Sea-Surface Temperature Anomalies",
    subtitle = paste0(
      "December 2009 and 2010 compared with the ",
      "1971–2000 climatological average"
    ),
    caption = "Source: NOAA Optimum Interpolation Sea Surface Temperature (OISST)"
  ) +

  theme_void() +
  theme(
    plot.title = element_text(
      face = "bold",
      size = 17
    ),
    plot.subtitle = element_text(
      size = 11,
      margin = margin(b = 10)
    ),
    plot.caption = element_text(
      size = 9,
      colour = "grey35"
    ),
    strip.text = element_text(
      face = "bold",
      size = 11,
      margin = margin(t = 4, b = 4)
    ),
    panel.spacing = unit(0.8, "lines"),
    legend.position = "right",
    legend.box = "vertical",
    legend.title = element_text(face = "bold"),
    plot.margin = margin(10, 10, 10, 10)
  )

binned_map
```

## 3.14 Reflect on the Visualization

### a. Difference between December 2009 El Niño and December 2010 La Niña

The December 2009 map shows a clear band of warm anomalies along the equatorial Pacific, especially in the central and eastern Pacific. In contrast, December 2010 shows a large area of cool anomalies across the same region. The 2009 map also contains stronger warm areas in parts of the southern Pacific, while the 2010 map has more widespread cooling across the tropical Pacific. This shows the opposite ocean temperature patterns associated with El Niño and La Niña.

### b. Gains and losses from showing only open ocean and grouping sea ice

Showing anomalies only over open ocean makes the temperature patterns easier to see because land and sea ice do not interfere with the colour scale. Grouping sea ice into one class also keeps the map and legend simple. However, the reader cannot see differences in sea-ice concentration above the 15% threshold. The map also does not show anomaly values beneath sea ice, so some information in the polar regions is hidden.

### c. Continuous scale compared with binned scale

The binned scale is better for comparing the general extent of warm and cool anomalies because the clear classes make similar regions easier to identify across both panels. It is also easier to tell which temperature range each colour represents. However, the binned scale removes small differences within each class and makes the map look less detailed. The continuous scale preserves gradual changes and fine spatial patterns, but readers need to estimate values from subtle colour differences.

# 4 Completing the Submission

## 4.1 Reflect on AI Use
Our team used Microsoft Copilot to support our understanding of the tasks and resolve coding issues. AI was mainly used to explain R functions, troubleshoot errors, and clarify geospatial concepts such as importing NetCDF files with read_stars(), assigning coordinate reference systems using st_set_crs(), checking spatial extents with st_bbox(), and understanding why file sorting and consistent anomaly colour scales are important. It also helped us understand the purpose of each step before we implemented the code ourselves. copilot_group_chat_transcript.pdf

We did not rely on AI to complete the lab for us. Instead, we reviewed its suggestions as a group and compared them with the lab instructions before deciding whether to use them. For example, AI initially suggested using packages such as terra, ncdf4, and tidync to read the NetCDF files, but after checking the assignment requirements, we followed the required approach using the stars package instead. Using AI in this way helped us improve our understanding while ensuring that the final report, code, and explanations reflected our own work and complied with the lab requirements.

The Copilot interaction log is provided in
[ai_log.pdf](ai_log.pdf).

## 4.2 Team Roles and Contributions

Jun Kiat served as the Coder and was responsible for creating and maintaining the main .qmd file, importing the datasets, implementing the code, and ensuring that the document rendered successfully.

Adam acted as the Navigator by guiding the team’s progress, interpreting the task requirements, and ensuring that all sections were completed.

Gerald took on the role of Verifier and carried out the required validation checks, including data verification, scale verification, and legend checks.

Bao Quan served as the Constructive Critic and Reporter by reviewing the team’s explanations, improving the clarity of written responses, and preparing explanations for the team’s plotting decisions.

Jia Jia acted as the AI Liaison by interacting with the chosen AI tool, maintaining the AI interaction log, and evaluating AI suggestions before they were incorporated into the submission.